### 2009-07-16 paper.tex
Initial IEEE paper shell with no content.

\documentclass[conference]{IEEEtran}

\newcommand{\thetitle}{On the Linear Structure of Network Traffic: Predicting Flow Behavior}

\input{packages}
\input{definitions}

\title{\vspace{-0.25em}\thetitle}
\author{
{\large{Stefan~Karpinski, John~R.~Gilbert, Elizabeth~M.~Belding}} \vspace{0.25em}\\
Department of Computer Science \\
University of California, Santa Barbara \vspace{0.35em}\\
\textit{\{sgk,gilbert,ebelding\}@cs.ucsb.edu}
}

\bibliographystyle{IEEEtran}

\newcommand{\figurename}{Figure}
\newcommand{\tablename}{Table}

\begin{document}
\maketitle

\bibliography{IEEE,references}

\end{document}

### 2009-07-16 paper.tex
A few initial paragraphs.

@@
 
+\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
+
+This paper is the first in a three-part series, employing non-parametric discrete mixture modeling to understand and analyze network traffic patterns at the level of individual flows and packets.\flownote
+All three papers employ the same theoretical framework but apply it to different aspects of network traffic analysis: prediction, classification and generation.
+In this paper we first develop the framework and validate it by demonstrating how it allows the prediction of the behaviors of individual flows from the observation of only a handful of packets.
+% The following papers will focus on classification of flows according to application type and realistic workload generation.
+
+The fundamental concept underlying this work is that of a \emph{linear representation of network traffic}~\cite{Karpinski08}.
+Linear representations operate on traffic traces at the level of flows.
+The representation function maps a set of flows to point in a vector space;
+the space will typically have many dimensions, and coordinates describe aspects of the  behavior of the flows.
+The function must satisfy a simple linearity condition which we will describe in detail later.
+% the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
+The complete behavior of a traffic trace may be expressed as a matrix, where each flow in the trace is described by a given row of the matrix.
+
+Karpinski~\emph{et~al.}~\cite{Karpinski08} observed that the standard simplifications used in network traffic modeling can be viewed as the multiplication of such traffic matrices by  low-rank matrices.
+They went on to demonstrate that such simplifications severely distort important behavioral properties of traffic traces, and to propose matrix factorization as an alternative and preferable means of model reduction.
+
+% A linear representation of traffic is a function in the space of all possible flows\flownote into a vector space, satisfying the following linearity condition:
+% \begin{quote}
+% The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
+% \end{quote}
+% The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
+
+% \begin{enumerate}
+%   \item each flow's behavior can be described by various properties, protocol type, port numbers, occurrences of packets sizes, and a sequence of inter-packet intervals.
+% \end{enumerate}
+
 \bibliography{IEEE,references}

### 2009-07-16 paper.tex
Special formatting instructions for Infocom 2010.
...

--- 2009-07-16 paper.tex
@@
 \documentclass[conference]{IEEEtran}
+\makeatletter
+\def\ps@headings{%
+\def\@oddhead{\mbox{}\scriptsize\rightmark \hfil \thepage}%
+\def\@evenhead{\scriptsize\thepage \hfil \leftmark\mbox{}}%
+\def\@oddfoot{}%
+\def\@evenfoot{}}
+\makeatother
+\pagestyle{headings}
 
@@
 
+\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
+
+This paper is the first in a three-part series, employing non-parametric discrete mixture modeling to understand and analyze network traffic patterns at the level of individual flows and packets.\flownote
+All three papers employ the same theoretical framework but apply it to different aspects of network traffic analysis: prediction, classification and generation.
+In this paper we present the framework and demonstrate how to predict detailed behavior of individual flows from the observation of only a handful of initial packets.
+% The following papers will focus on classification of flows according to application type and realistic workload generation.
+
+The fundamental concept underlying this work is that of a \emph{linear representation of network traffic}~\cite{Karpinski08}.
+Linear representations operate on traffic traces at the level of flows.
+A representation is a function that maps sets of flows to points in a vector space;
+the space will typically have high dimension, each coordinate describing some aspect of flow behavior.
+The representation must satisfy a simple linearity condition which we will describe in detail later.
+% the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
+The complete behavior of a traffic trace is expressed as a matrix, where each flow in the trace is represented by a row of the matrix.
+
+% Karpinski~\emph{et~al.}~\cite{Karpinski08} further observed that the simplifications commonly used in traffic modeling correspond to multiplication of traffic representation matrices by low-rank matrices, averaging over rows or columns.
+% They demonstrated experimentally that such simplifications severely distort important behavioral properties of traffic traces;
+% viewing these simplification as low-rank matrix transformations, this is unsurprising:
+% most of the detail of the original traffic behavior is destroyed.
+% It was suggested that matrix factorization provides an alternative and superior means of model reduction.
+
+The key observation of this work is that even real network traces tend to have approximately low-rank representation matrices.
+
 \bibliography{IEEE,references}

### 2009-07-17 paper.tex
...

@@
 This paper is the first in a three-part series, employing non-parametric discrete mixture modeling to understand and analyze network traffic patterns at the level of individual flows and packets.\flownote
-All three papers employ the same theoretical framework but apply it to different aspects of network traffic analysis: prediction, classification and generation.
-In this paper we present the framework and demonstrate how to predict detailed behavior of individual flows from the observation of only a handful of initial packets.
+All three papers employ the same theoretical framework but investigate different fundamental applications of traffic modeling:
+prediction, classification and generation.
+In this paper we first present the framework and then demonstrate how it can predict detailed behavior of individual flows from the observation of only a handful of initial packets.
 % The following papers will focus on classification of flows according to application type and realistic workload generation.
 
-The fundamental concept underlying this work is that of a \emph{linear representation of network traffic}~\cite{Karpinski08}.
-Linear representations operate on traffic traces at the level of flows.
-A representation is a function that maps sets of flows to points in a vector space;
-the space will typically have high dimension, each coordinate describing some aspect of flow behavior.
-The representation must satisfy a simple linearity condition which we will describe in detail later.
+The fundamental concept underlying these works is \emph{linear representation of network traffic}~\cite{Karpinski08}.
+A linear representation is a function that maps sets of flows to points in a vector space;
+each dimension of the representation describes some aspect of flow behavior.
+The representation must also satisfy a simple linearity condition which we will describe in detail later.
 % the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
-The complete behavior of a traffic trace is expressed as a matrix, where each flow in the trace is represented by a row of the matrix.
+An entire traffic trace may be represented as a matrix:
+each flow is mapped to a row of the matrix.
 
@@
 
-The key observation of this work is that even real network traces tend to have approximately low-rank representation matrices.
+The key observation of this work is that real-world network traces tend to have approximately low-rank representation matrices.
 

### 2009-07-17 paper.tex
...

@@
 % the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
-An entire traffic trace may be represented as a matrix:
-each flow is mapped to a row of the matrix.
+An entire traffic trace may be represented as a matrix where each row is the linear representation of a flow in the trace.
+We restrict our consideration to a particularly natural and powerful class of representations whose coordinates describe the frequencies of flow features.
+For example, let there be a dimension for each packet size up to 1500.
+Each coordinate indicates the number of packets in a set of flows of that size.
+\caps{IP} protocol type, port numbers, and inter-packet intervals can be handled similarly.
+% Examples of flow features are the number of packets of a certain size or the use of a particular port number.
+% The fundamental premise of this work, and of the sequel, is that the representatives of actual flow behaviors can be understood as being sampled from
+
+Our key observation is that real-world network traces tend to have approximately low-rank representation matrices despite the potential rank of such matrices being very large.
+Moreover, most flows' vectors lie on only one of a few even lower rank subspaces.
+Real traffic lies near the union of a small set of very low rank subspaces in the huge potential space of representations.
+Moreover, the low-rank subspace near which real traffic lies can be decomposed into subspaces of even lower rank, with almost all flows lying in one of these subspaces.
+Typically, these smaller subspaces are one-dimensional, but in some cases they may have up to a dozen dimensions.
+
+% the vast majority of observed flow behaviors in any given trace lie in one of a handful low-rank subspaces.
+% These subspaces are mostly of rank one, but a few have dimensions as high as ten.
 
@@
 
-The key observation of this work is that real-world network traces tend to have approximately low-rank representation matrices.
-
 \bibliography{IEEE,references}

### 2009-07-17 paper.tex
...

@@
 
-Our key observation is that real-world network traces tend to have approximately low-rank representation matrices despite the potential rank of such matrices being very large.
-Moreover, most flows' vectors lie on only one of a few even lower rank subspaces.
-Real traffic lies near the union of a small set of very low rank subspaces in the huge potential space of representations.
-Moreover, the low-rank subspace near which real traffic lies can be decomposed into subspaces of even lower rank, with almost all flows lying in one of these subspaces.
-Typically, these smaller subspaces are one-dimensional, but in some cases they may have up to a dozen dimensions.
+The essential observation of this research is that real traffic lies near the union of a small set of low rank subspaces in the representation space.
+This allows us to perform a drastic model reduction on traffic data while preserving most characteristics of the original traffic.
+Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression, and allows us to very simply predict unseen behavior of flows.
+We apply this technique here to directly predict the following characteristics of each flow from observing only a handful of its initial packets:
+\begin{enumerate}
+  \item the distribution of packets sizes,
+  \item the distribution of inter-packet intervals,
+  \item the number of packets that will be in the flow.
+\end{enumerate}
+From these properties, we can indirectly predict many other properties, including the total duration and data volume.
+% 
+% real-world network traces tend to have approximately low-rank representation matrices despite the potential rank of such matrices being very large.
+% Moreover, most flows' vectors lie on only one of a few even lower rank subspaces.
+% Moreover, the low-rank subspace near which real traffic lies can be decomposed into subspaces of even lower rank, with almost all flows lying in one of these subspaces.
+% Typically, these smaller subspaces are one-dimensional, but in some cases they may have up to a dozen dimensions.
 

### 2009-07-17 paper.tex
...

@@
 
-The fundamental concept underlying these works is \emph{linear representation of network traffic}~\cite{Karpinski08}.
-A linear representation is a function that maps sets of flows to points in a vector space;
-each dimension of the representation describes some aspect of flow behavior.
-The representation must also satisfy a simple linearity condition which we will describe in detail later.
+The fundamental concept underlying this work is \emph{linear representation of network traffic}~\cite{Karpinski08}.
+A linear representation is a function that maps sets of flows to vectors, the coordinates of which describe aspects of flow behavior;
+the representation must also satisfy a simple linearity condition which we will describe in detail later.
 % the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
-An entire traffic trace may be represented as a matrix where each row is the linear representation of a flow in the trace.
+An entire traffic trace may be represented as a matrix whose rows are the representations of flows in the trace.
 We restrict our consideration to a particularly natural and powerful class of representations whose coordinates describe the frequencies of flow features.
-For example, let there be a dimension for each packet size up to 1500.
-Each coordinate indicates the number of packets in a set of flows of that size.
-\caps{IP} protocol type, port numbers, and inter-packet intervals can be handled similarly.
+For example, the representation can have a dimension for each packet size up to 1500; coordinates in these dimensions indicate the number of packets of each size.
+Port numbers, \caps{IP} protocol type, and inter-packet intervals can be handled similarly.
 % Examples of flow features are the number of packets of a certain size or the use of a particular port number.
@@
 \end{enumerate}
-From these properties, we can indirectly predict many other properties, including the total duration and data volume.
+From these properties we can indirectly predict many other properties, including the total duration and data volume.
 % 

### 2009-07-17 paper.tex
...

@@
 
-The essential observation of this research is that real traffic lies near the union of a small set of low rank subspaces in the representation space.
-This allows us to perform a drastic model reduction on traffic data while preserving most characteristics of the original traffic.
-Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression, and allows us to very simply predict unseen behavior of flows.
-We apply this technique here to directly predict the following characteristics of each flow from observing only a handful of its initial packets:
+The essential observation of this research is that under such representations, real flows lie near the union of a small set of low rank subspaces.
+This property allows us to perform a drastic model reduction on trace traffic patterns while preserving most characteristics of the original traffic.
+Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:
+it allows us to very simply predict unseen behavior of flows.
+We use this technique to predict the following characteristics of each flow from observing only a handful of its initial packets:
 \begin{enumerate}
@@
 From these properties we can indirectly predict many other properties, including the total duration and data volume.
-% 
+
+At this point it may be wise to address what we mean by prediction.
+We do not attempt to predict the exact number of packets in a specific flow.
+Brief reflection shows that this cannot be done accurately:
+flow behavior is not deterministic---many flows exhibit the same behavior initially but subsequently behave quite differently.
+The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+
 % real-world network traces tend to have approximately low-rank representation matrices despite the potential rank of such matrices being very large.

### 2009-07-17 paper.tex
...

@@
 At this point it may be wise to address what we mean by prediction.
-We do not attempt to predict the exact number of packets in a specific flow.
-Brief reflection shows that this cannot be done accurately:
-flow behavior is not deterministic---many flows exhibit the same behavior initially but subsequently behave quite differently.
+Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
+flow behavior is nondeterministic---flows may exhibit the same initial behavior but subsequently behave quite differently.
 The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.

### 2009-07-17 paper.tex
...

@@
 
-\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}
+\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': IP protocol type, source and destination nodes, and TCP/UDP port numbers.}
 
@@
 The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+When compared with a specific outcome, such a prediction cannot be said to be ``wrong or right.''
+How then can one measure the quality of a predicted distribution of outcomes?
+This is a subtle and difficult issue is central to our evaluation methodology.
 

### 2009-07-17 paper.tex
...

@@
 Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:
-it allows us to very simply predict unseen behavior of flows.
+it allows us to use a least squares optimization on observed features to predict behavior.
 We use this technique to predict the following characteristics of each flow from observing only a handful of its initial packets:
@@
 The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-When compared with a specific outcome, such a prediction cannot be said to be ``wrong or right.''
+When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
 How then can one measure the quality of a predicted distribution of outcomes?
-This is a subtle and difficult issue is central to our evaluation methodology.
+This subtle and difficult issue is central to our evaluation methodology.
 

### 2009-07-21 paper.tex
...

@@
 A linear representation is a function that maps sets of flows to vectors, the coordinates of which describe aspects of flow behavior;
-the representation must also satisfy a simple linearity condition which we will describe in detail later.
+it must distribute additively over set unions.
+% the representation must also satisfy a simple linearity condition which we will describe in detail later.
 % the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
 An entire traffic trace may be represented as a matrix whose rows are the representations of flows in the trace.
-We restrict our consideration to a particularly natural and powerful class of representations whose coordinates describe the frequencies of flow features.
+We restrict our consideration to a particularly natural and powerful class of representations whose coordinates are the frequencies of flow features in a set of flows.
 For example, the representation can have a dimension for each packet size up to 1500; coordinates in these dimensions indicate the number of packets of each size.
@@
 
-The essential observation of this research is that under such representations, real flows lie near the union of a small set of low rank subspaces.
-This property allows us to perform a drastic model reduction on trace traffic patterns while preserving most characteristics of the original traffic.
+The essential observation of this research is that under such representations, real flows lie near the union of a small set of low-rank subspaces.
+This property allows us to perform a drastic model reduction on trace traffic patterns while preserving the essential characteristics of the original traffic.
 Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:

### 2009-07-22 paper.tex
...

@@
 
+\section{Introduction}
+
 \newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': IP protocol type, source and destination nodes, and TCP/UDP port numbers.}
@@
 
-The essential observation of this research is that under such representations, real flows lie near the union of a small set of low-rank subspaces.
+The essential observation of this research is that under such representations, real flows tend to lie near a small set of low-rank subspaces.
 This property allows us to perform a drastic model reduction on trace traffic patterns while preserving the essential characteristics of the original traffic.
@@
 
+\section{Background \& Motivation}
+
+The most fundamental tool in the investigation of linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
+This matrix decomposition can be applied to any real matrix and factors it into a product of three matrices:
+\begin{align}
+  X = US\trans{V}
+\end{align}
+where $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
+Among the useful properties of this factorization, if all but the first $k$ diagonal entries of $S$ are zeroed out, the product $US_k\trans{V}$ is the best rank-$k$ approximation of $X$ under the Frobenius matrix norm.
+The rows of $US_k$ can thus be viewed as an optimal $k$-dimensional simplification of the rows of $X$ in the new coordinate system given by the columns of $V$.
+
+\begin{figure}[!b]
+\vspace{0.5em}
+\begin{center}
+\subfloat[IETF 60]{\includegraphics[width=1.6in]{svd/ie60}}
+\subfloat[IETF 67]{\includegraphics[width=1.6in]{svd/ie67}}
+\subfloat[SIGCOMM 2004]{\includegraphics[width=1.6in]{svd/sc04}}
+\subfloat[UCSD]{\includegraphics[width=1.6in]{svd/ucsd}}
+\caption{SVD scatter plots for various traces.} 
+\fig{svd}
+\end{center}
+\vspace{-0.85em}
+\end{figure}
+
+% The most direct way to demonstrate the motivating insight of this research is with pictures.
+
+\section{Methodology}
+
+\section{Results}
+
+\section{Discussion}
+
+\section{Conclusions}
+
 \bibliography{IEEE,references}

### 2009-07-22 paper.tex
...

@@
 
-\section{Introduction}
+\section{Introduction}\sec{introduction}
 
@@
 
-\section{Background \& Motivation}
+\section{Background \& Motivation}\sec{background}\sec{motivation}
 
-The most fundamental tool in the investigation of linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
-This matrix decomposition can be applied to any real matrix and factors it into a product of three matrices:
+The most fundamental tool for investigating linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
+This matrix decomposition can be applied to any real matrix and factors it into a product of three real matrices:
 \begin{align}
-  X = US\trans{V}
+  X = US\trans{V}.
 \end{align}
-where $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
+Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
 Among the useful properties of this factorization, if all but the first $k$ diagonal entries of $S$ are zeroed out, the product $US_k\trans{V}$ is the best rank-$k$ approximation of $X$ under the Frobenius matrix norm.
-The rows of $US_k$ can thus be viewed as an optimal $k$-dimensional simplification of the rows of $X$ in the new coordinate system given by the columns of $V$.
+The rows of $US_k$ can thus be viewed as an optimal $k$-dimensional reduced model of the rows of $X$, with a new coordinate system given by the columns of $V$.
 
-\begin{figure}[!b]
+We can visualize the $k$-dimensional approximate structure of $X$ in $k$ dimensions using the coordinates of $US_k$.
+However, the most significant feature of any vector in this representation is its magnitude.
+We are more interested in the direction vectors than their direction:
+points with the same direction lie in the same one-dimensional subspace.
+To visualize direction while eliminating distance, we project all data points radially onto the hyperplane in which all points have coordinates summing to unity.\footnote{We could normalize the vectors to unit length, but then our projection would be spherical, artificially introducing curvature into our visualization.}
+% Then we simply scatter-plot various pairs of dimensions of the resulting data.
+\Figure{svd} shows scatter plots of the most significant two projected \caps{SVD} coordinates for four different traces.
+The original coordinates of this data encode the packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows for each trace.
+The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
+
+\begin{figure}[!t]
 \vspace{0.5em}
@@
 
-\section{Methodology}
+\section{Methodology}\sec{methodology}
 
-\section{Results}
+\section{Results}\sec{results}
 
-\section{Discussion}
+\section{Discussion}\sec{discussion}
 
-\section{Conclusions}
+\section{Conclusions}\sec{conclusions}
 

### 2009-07-22 paper.tex
...

@@
 \begin{center}
-\subfloat[IETF 60]{\includegraphics[width=1.6in]{svd/ie60}}
-\subfloat[IETF 67]{\includegraphics[width=1.6in]{svd/ie67}}
-\subfloat[SIGCOMM 2004]{\includegraphics[width=1.6in]{svd/sc04}}
-\subfloat[UCSD]{\includegraphics[width=1.6in]{svd/ucsd}}
+\subfloat[\footnotesize{IETF 60}]{\includegraphics[width=1.6in]{svd/ie60}}
+\subfloat[\footnotesize{IETF 67}]{\includegraphics[width=1.6in]{svd/ie67}}
+\subfloat[\footnotesize{SIGCOMM 2004}]{\includegraphics[width=1.6in]{svd/sc04}}
+\subfloat[\footnotesize{UCSD}]{\includegraphics[width=1.6in]{svd/ucsd}}
 \caption{SVD scatter plots for various traces.} 
@@
 
-% The most direct way to demonstrate the motivating insight of this research is with pictures.
+Clearly visible in \Figure{svd} is linear structure:
+points are heavily concentrated on low-dimensional subspaces.
+Because of the projection step of our process, points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
+points concentrated along a two-dimensional subspace appear as lines;
+points falling along subspaces of three or more dimensions appear as ``smears.''
 

### 2009-07-23 paper.tex
...

@@
 
-At this point it may be wise to address what we mean by prediction.
+At this point, it may be wise to address what we mean here by prediction.
 Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
@@
 
-The most fundamental tool for investigating linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
+The foremost tool for investigating linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
 This matrix decomposition can be applied to any real matrix and factors it into a product of three real matrices:
@@
 Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
-Among the useful properties of this factorization, if all but the first $k$ diagonal entries of $S$ are zeroed out, the product $US_k\trans{V}$ is the best rank-$k$ approximation of $X$ under the Frobenius matrix norm.
-The rows of $US_k$ can thus be viewed as an optimal $k$-dimensional reduced model of the rows of $X$, with a new coordinate system given by the columns of $V$.
+Among the useful properties of this factorization, is that the product $US_k\trans{V}$ is the best rank-$k$ approximation of $X$ under the Frobenius matrix norm, where $S_k$ is the matrix $S$ with all but the first $k$ diagonal entries zeroed out.
+The $k$ initial coordinates of the rows of $US_k$ can thus be viewed as an optimal $k$-dimensional reduced model of the rows of $X$, with a new coordinate system given by the columns of $V$.
 
@@
 \end{center}
-\vspace{-0.85em}
+\vspace{-2em}
 \end{figure}
 
-Clearly visible in \Figure{svd} is linear structure:
-points are heavily concentrated on low-dimensional subspaces.
+Linear structure is clearly visible in \Figure{svd}:
+the points are heavily concentrated on low-dimensional subspaces.
 Because of the projection step of our process, points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;

### 2009-07-23 paper.tex
...

@@
 
+\newfootnote{\svdnote}{More precisely, the product of the first $k$ columns of $U$ with the first $k$ rows of $\trans{V}$ is a Frobenius-optimal rank-$k$ approximation of $X$.}
+
+\newfootnote{\projectionnote}{We could normalize the vectors to unit length, but then our projection would be spherical, artificially introducing curvature into our visualization.}
+
 The foremost tool for investigating linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
@@
 Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
-Among the useful properties of this factorization, is that the product $US_k\trans{V}$ is the best rank-$k$ approximation of $X$ under the Frobenius matrix norm, where $S_k$ is the matrix $S$ with all but the first $k$ diagonal entries zeroed out.
-The $k$ initial coordinates of the rows of $US_k$ can thus be viewed as an optimal $k$-dimensional reduced model of the rows of $X$, with a new coordinate system given by the columns of $V$.
+Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of the rows of $X$, in a new coordinate system given by the first $k$ columns of $V$\!.\,\svdnote
 
-We can visualize the $k$-dimensional approximate structure of $X$ in $k$ dimensions using the coordinates of $US_k$.
-However, the most significant feature of any vector in this representation is its magnitude.
-We are more interested in the direction vectors than their direction:
+This property allows us to visualize the approximate structure of $X$ using initial columns of $US$.
+However, the most significant feature of any vector in this representation is its magnitude, while we are interested in the direction of vectors:
 points with the same direction lie in the same one-dimensional subspace.
-To visualize direction while eliminating distance, we project all data points radially onto the hyperplane in which all points have coordinates summing to unity.\footnote{We could normalize the vectors to unit length, but then our projection would be spherical, artificially introducing curvature into our visualization.}
-% Then we simply scatter-plot various pairs of dimensions of the resulting data.
+To visualize direction while eliminating distance, we project all data points radially onto the hyperplane in which all points have coordinates summing to unity.\projectionnote
 \Figure{svd} shows scatter plots of the most significant two projected \caps{SVD} coordinates for four different traces.

### 2009-07-23 paper.tex
...

@@
 This property allows us to visualize the approximate structure of $X$ using initial columns of $US$.
-However, the most significant feature of any vector in this representation is its magnitude, while we are interested in the direction of vectors:
-points with the same direction lie in the same one-dimensional subspace.
-To visualize direction while eliminating distance, we project all data points radially onto the hyperplane in which all points have coordinates summing to unity.\projectionnote
-\Figure{svd} shows scatter plots of the most significant two projected \caps{SVD} coordinates for four different traces.
-The original coordinates of this data encode the packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows for each trace.
+We are, however, for the moment only interested in direction not magnitude:
+vectors with the same direction lie in the same one-dimensional subspace.
+To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane through the origin.\projectionnote
+\Figure{svd} shows scatter plots of the most significant two projected \caps{SVD} coordinates for four different traces of network traffic.
+The original data encodes packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows from each trace.
 The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
@@
 \subfloat[\footnotesize{UCSD}]{\includegraphics[width=1.6in]{svd/ucsd}}
-\caption{SVD scatter plots for various traces.} 
+\caption{SVD scatter plots for various traffic traces.} 
 \fig{svd}
@@
 points concentrated along a two-dimensional subspace appear as lines;
-points falling along subspaces of three or more dimensions appear as ``smears.''
+points falling near subspaces of more dimensions appear as ``smears.''
 

### 2009-07-23 paper.tex
...

@@
 
+\begin{figure*}[t]
+\begin{center}
+\subfloat[\footnotesize{DARTMOUTH}]{\includegraphics[width=1.15in]{svd/dart}}
+\subfloat[\footnotesize{IETF 60}]{\includegraphics[width=1.15in]{svd/ie60}}
+\subfloat[\footnotesize{IETF 67}]{\includegraphics[width=1.15in]{svd/ie67}}
+\subfloat[\footnotesize{SIGCOMM 2001}]{\includegraphics[width=1.15in]{svd/sc01}}
+\subfloat[\footnotesize{SIGCOMM 2004}]{\includegraphics[width=1.15in]{svd/sc04}}
+\subfloat[\footnotesize{UCSD}]{\includegraphics[width=1.15in]{svd/ucsd}}
+\caption{SVD flow-behavior scatter plots for the six network traffic traces considered in this paper.} 
+\fig{svd}
+\end{center}
+\vspace{-2em}
+\end{figure*}
+
+A linear representation, as defined by Karpinski~\emph{et~al.}~\cite{Karpinski08}, maps sets of flows into a vector space such that union of sets becomes summation in the vector space:
+\begin{align}
+  \Phi(s_1 \union s_2) = \Phi(s_1) + \Phi(s_2).
+\end{align}
+The representation function, $\Phi$, should map each set of flows to a vector which in some sense describes their collective behavior.
+For example: let $\Phi(s) \in \R^1$ be the number of packets in all the flows in as set $s$.
+Since the number of total packets in the union of two sets of flows is the sum of the number in each individually, this representation is linear.
+
 \newfootnote{\svdnote}{More precisely, the product of the first $k$ columns of $U$ with the first $k$ rows of $\trans{V}$ is a Frobenius-optimal rank-$k$ approximation of $X$.}
@@
 
-\begin{figure}[!t]
-\vspace{0.5em}
-\begin{center}
-\subfloat[\footnotesize{IETF 60}]{\includegraphics[width=1.6in]{svd/ie60}}
-\subfloat[\footnotesize{IETF 67}]{\includegraphics[width=1.6in]{svd/ie67}}
-\subfloat[\footnotesize{SIGCOMM 2004}]{\includegraphics[width=1.6in]{svd/sc04}}
-\subfloat[\footnotesize{UCSD}]{\includegraphics[width=1.6in]{svd/ucsd}}
-\caption{SVD scatter plots for various traffic traces.} 
-\fig{svd}
-\end{center}
-\vspace{-2em}
-\end{figure}
-
 Linear structure is clearly visible in \Figure{svd}:
@@
 points falling near subspaces of more dimensions appear as ``smears.''
+% There is a smattering of outliers, not especially near such structures.
 

### 2009-07-23 paper.tex
...

@@
 We restrict our consideration to a particularly natural and powerful class of representations whose coordinates are the frequencies of flow features in a set of flows.
-For example, the representation can have a dimension for each packet size up to 1500; coordinates in these dimensions indicate the number of packets of each size.
+For example, the representation can have a dimension for each packet size up to the maximum transfer unit (\caps{MTU});
+coordinates in these dimensions indicate the count of packets of each size.
 Port numbers, \caps{IP} protocol type, and inter-packet intervals can be handled similarly.
@@
 The representation function, $\Phi$, should map each set of flows to a vector which in some sense describes their collective behavior.
-For example: let $\Phi(s) \in \R^1$ be the number of packets in all the flows in as set $s$.
+For example: let $\Phi(s) \in \R^1$ be the number of packets in all the flows in the set $s$.
 Since the number of total packets in the union of two sets of flows is the sum of the number in each individually, this representation is linear.
+Given two representations, their direct sum is readily seen to also be a linear representation:
+\begin{align}
+  \Phi(s) = \Phi_1(s) \directsum \Phi_2(s).
+\end{align}
+
+A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
+As mentioned before, a 1500-dimensional vector can be used to count the frequency of all possible packets sizes from a single byte up to the typical \caps{MTU} of 1500 bytes.
+In this representation, $\Phi(s) \in \R^{1500}$ has as its $i$th entry the number of packets in $s$ of size $i$.
+Since \caps{IP} protocol types range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
+The representation details for port numbers and inter-packet intervals are described detail in \Section{methodology}.
+% Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
+% For practical purposes, however, this is too large a vector space, so we represent port numbers in 1000-dimensions, and reduce port numbers modulo 1000.
+% Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
+The representation we consider here is a direct sum of such histogram representations.
+The combination is a representation in 5256 dimensions.
+An entire trace is represented as a matrix of such representation vectors---one flow per row.
+This paper explores the linear structure of such matrices.
 

### 2009-07-23 paper.tex
...

@@
 
-A linear representation, as defined by Karpinski~\emph{et~al.}~\cite{Karpinski08}, maps sets of flows into a vector space such that union of sets becomes summation in the vector space:
+A linear representation, as proposed by Karpinski~\emph{et~al.}~\cite{Karpinski08}, maps sets of flows into a vector space such that disjoint union of sets becomes addition in the vector space:
 \begin{align}
-  \Phi(s_1 \union s_2) = \Phi(s_1) + \Phi(s_2).
+  \Phi(s_1 \disunion s_2) = \Phi(s_1) + \Phi(s_2).
 \end{align}
-The representation function, $\Phi$, should map each set of flows to a vector which in some sense describes their collective behavior.
-For example: let $\Phi(s) \in \R^1$ be the number of packets in all the flows in the set $s$.
-Since the number of total packets in the union of two sets of flows is the sum of the number in each individually, this representation is linear.
+The representation function should map each set of flows to a vector that in some sense describes their collective behavior.
+For example: let $\Phi(s) \in \R^1$ be the number of packets in the flows in $s$.
+Since the number of packets in the disjoint union of sets of flows is the sum of the number in each set individually, this representation is linear.
 Given two representations, their direct sum is readily seen to also be a linear representation:
@@
 A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
-As mentioned before, a 1500-dimensional vector can be used to count the frequency of all possible packets sizes from a single byte up to the typical \caps{MTU} of 1500 bytes.
-In this representation, $\Phi(s) \in \R^{1500}$ has as its $i$th entry the number of packets in $s$ of size $i$.
-Since \caps{IP} protocol types range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
-The representation details for port numbers and inter-packet intervals are described detail in \Section{methodology}.
+A 1500-dimensional vector can be used to count the frequency of packet sizes seen, from one byte up to the typical \caps{MTU} of 1500 bytes.
+In this representation, $\Phi(s)$ has as its $i$th entry the number of packets in $s$ of size $i$.
+Since \caps{IP} protocols range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
+Histogram representations for port numbers and inter-packet intervals are detailed in \Section{methodology}.
 % Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
@@
 % Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
-The representation we consider here is a direct sum of such histogram representations.
+The representation we use is a direct sum of these histogram representations.
 The combination is a representation in 5256 dimensions.
@@
 
-The foremost tool for investigating linear structure in high dimensions is the singular value decomposition (\caps{SVD}).
-This matrix decomposition can be applied to any real matrix and factors it into a product of three real matrices:
+To see the linear structure of these representation matrices, we apply one of the most powerful tools of linear algebra:
+the singular value decomposition (\caps{SVD}).
+This matrix decomposition can be applied to any real matrix, factoring it into a product of three real matrices:
 \begin{align}
@@
 \end{align}
-Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of decreasing nonnegative value.
-Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of the rows of $X$, in a new coordinate system given by the first $k$ columns of $V$\!.\,\svdnote
+Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
+Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\,\svdnote
+This property allows us to visualize the structure of $X$ using initial columns of $US$.
+However, we are interested only interested in direction since
+vectors differing only by magnitude lie in the same one-dimensional subspace.
+To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane through the origin.\projectionnote
 
-This property allows us to visualize the approximate structure of $X$ using initial columns of $US$.
-We are, however, for the moment only interested in direction not magnitude:
-vectors with the same direction lie in the same one-dimensional subspace.
-To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane through the origin.\projectionnote
-\Figure{svd} shows scatter plots of the most significant two projected \caps{SVD} coordinates for four different traces of network traffic.
-The original data encodes packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows from each trace.
-The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
-
-Linear structure is clearly visible in \Figure{svd}:
-the points are heavily concentrated on low-dimensional subspaces.
-Because of the projection step of our process, points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
+\Figure{svd} illustrates the results of applying \caps{SVD} dimension reduction to sample matrices of six different traffic traces.
+The figures show scatter plots of the most significant two projected \caps{SVD} coordinates for a matrix of 5000 sample flows from each trace.
+% The original data encodes packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows from each trace.
+% The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
+Linear structure is clearly visible in every one:
+flow behaviors are heavily concentrated near low-dimensional subspaces.
+Points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
 points concentrated along a two-dimensional subspace appear as lines;

### 2009-07-23 paper.tex
...

@@
 \end{align}
-Here $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
+$U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
 Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\,\svdnote
@@
 % The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
-Linear structure is clearly visible in every one:
+Linear structure is visible in all six:
 flow behaviors are heavily concentrated near low-dimensional subspaces.
 Points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
-points concentrated along a two-dimensional subspace appear as lines;
-points falling near subspaces of more dimensions appear as ``smears.''
+points concentrated near two-dimensional subspaces appear as lines;
+points falling along higher subspaces appear as ``smears.''
 % There is a smattering of outliers, not especially near such structures.

### 2009-07-23 paper.tex
...

@@
 \begin{align}
-  X = US\trans{V}.
+  X = U\!S\trans{V}.
 \end{align}
 $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
-Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\,\svdnote
-This property allows us to visualize the structure of $X$ using initial columns of $US$.
+Among the useful properties of this factorization, is that the first $k$ columns of $U\!S$ are an optimal $k$-dimensional reduced model of $X$.\,\svdnote
+This property allows us to visualize the structure of $X$ using initial columns of $U\!S$.
 However, we are interested only interested in direction since

### 2009-07-23 paper.tex
...

@@
 $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
-Among the useful properties of this factorization, is that the first $k$ columns of $U\!S$ are an optimal $k$-dimensional reduced model of $X$.\,\svdnote
+Among the useful properties of this factorization, is that the first $k$ columns of $U\!S$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
 This property allows us to visualize the structure of $X$ using initial columns of $U\!S$.
@@
 
+The presence of linear structure in the matrix representations of several independent traffic traces immediately raises two significant questions.
+The more philosophical question is \emph{Why does this linear structure exist?}
+The more practical question is \emph{How can we exploit this structure?}
+Neither question can be answered fully in a single paper.
+In what follows, we provide some insight into the first question, and use that insight to predict the behavior of individual flows from the observation of only a few initial packets.
+
+\vfill
+\break
+
 \section{Methodology}\sec{methodology}

### 2009-07-24 paper.tex
...

@@
 vectors differing only by magnitude lie in the same one-dimensional subspace.
-To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane through the origin.\projectionnote
+To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane from the origin.\projectionnote
 
@@
 flow behaviors are heavily concentrated near low-dimensional subspaces.
-Points concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
-points concentrated near two-dimensional subspaces appear as lines;
-points falling along higher subspaces appear as ``smears.''
+Behaviors concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
+behaviors near two-dimensional subspaces appear as lines;
+behaviors on higher-dimensional subspaces appear as ``smears.''
 % There is a smattering of outliers, not especially near such structures.
@@
 The presence of linear structure in the matrix representations of several independent traffic traces immediately raises two significant questions.
-The more philosophical question is \emph{Why does this linear structure exist?}
-The more practical question is \emph{How can we exploit this structure?}
-Neither question can be answered fully in a single paper.
-In what follows, we provide some insight into the first question, and use that insight to predict the behavior of individual flows from the observation of only a few initial packets.
+The existential question: \emph{Why does this linear structure exist?}
+Followed, of course, by the practical question: \emph{How can we exploit this structure?}
+Neither question can be answered fully in a single paper, but we will begin to address the existential problem now.
+In the rest of the paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
 
@@
 \section{Methodology}\sec{methodology}
+
+\subsection{Linear Representation}\sec{linear-representation}
+
+\subsection{Subspace Segmentation}\sec{subspace-segmentation}
+
+\subsection{Subspace Factorization}\sec{subspace-factorization}
+
+\subsection{Flow Behavior Prediction}\sec{flow-behavior-prediction}
 

### 2009-07-24 paper.tex
...

@@
 Followed, of course, by the practical question: \emph{How can we exploit this structure?}
-Neither question can be answered fully in a single paper, but we will begin to address the existential problem now.
+Neither question can be answered fully in a single paper, but we will begin to address the existential problem in the next section.
 In the rest of the paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
+
+\section{Whence Linearity?}
+
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be posed in the form of a model.
+A mixture model, to be specific.
+
+% The question of why we see low-dimensional linear structure in the feature-frequency representation of network traffic is deeply connected to the nature of such linear representations themselves.
+% But the answer to the question also requires us to posit a hypothesis about the nature of network traffic.
 

### 2009-07-24 paper.tex
...

@@
 A mixture model, to be specific.
+We propose to view each flow as having a probability distribution of possible behaviors it could exhibit.
+% We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
+We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
+Moreover, only even smaller subsets of these basic behaviors are ever mixed with each other.
+
+% Mixture modeling expresses probability densities as convex combinations of constituent probability distributions:
+% \begin{align}\eqn{mixture-model}
+%   q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
+% \end{align}
+% Here $q_i$ and $p_j$ are density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
+% \Equation{mixture-model} says that each of the distributions $q_i$ is a convex mixture of the
 

### 2009-07-24 paper.tex
...

@@
 Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be posed in the form of a model.
-A mixture model, to be specific.
-We propose to view each flow as having a probability distribution of possible behaviors it could exhibit.
+% A mixture model, to be specific.
+We propose viewing each flow as having a probability distribution of possible behaviors it could exhibit.
 % We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
@@
 Moreover, only even smaller subsets of these basic behaviors are ever mixed with each other.
+Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
+\begin{align}\eqn{mixture-model}
+  q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
+\end{align}
+Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
 

### 2009-07-24 paper.tex
...

@@
 
-This paper is the first in a three-part series, employing non-parametric discrete mixture modeling to understand and analyze network traffic patterns at the level of individual flows and packets.\flownote
+This paper is the first in a three-part series employing a numerical linear algebra techniques to understand and analyze network traffic patterns at a detailed level of individual flows and packets.\flownote
 All three papers employ the same theoretical framework but investigate different fundamental applications of traffic modeling:
 prediction, classification and generation.
-In this paper we first present the framework and then demonstrate how it can predict detailed behavior of individual flows from the observation of only a handful of initial packets.
+In this paper we first present the framework and then demonstrate how it can predict the behavior of individual flows from the observation of only a handful of their initial packets.
 % The following papers will focus on classification of flows according to application type and realistic workload generation.

### 2009-07-24 paper.tex
...

@@
 The representation we use is a direct sum of these histogram representations.
-The combination is a representation in 5256 dimensions.
+% The combination is a representation in 5256 dimensions.
 An entire trace is represented as a matrix of such representation vectors---one flow per row.
@@
 \begin{align}
-  X = U\!S\trans{V}.
+  X = US\trans{V}.
 \end{align}
 $U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
-Among the useful properties of this factorization, is that the first $k$ columns of $U\!S$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
-This property allows us to visualize the structure of $X$ using initial columns of $U\!S$.
+Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
+This property allows us to visualize the structure of $X$ using initial columns of $US$.
 However, we are interested only interested in direction since
@@
 
-\section{Whence Linearity?}
+\section{Model \& Hypothesis}
 
-Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be posed in the form of a model.
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a stochastic model of flow behavior.
 % A mixture model, to be specific.
@@
 We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
-Moreover, only even smaller subsets of these basic behaviors are ever mixed with each other.
+Moreover, only even smaller subsets of these basic behaviors are typically combined with each other.
 Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
@@
 Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
+\Equation{mixture-model} is expressed succinctly as matrix multiplication.
+Writing $Q_{ik} = q_i(k)$, $W_{ij} = w_{ij}$, and $P_{jk} = p_j(k)$, we have:
+\begin{align}\eqn{mixture-model-matrix}
+  Q = WP.
+\end{align}
+The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
+Next, partition the rows of $P$ into small sets which can have simultaneously non-zero weights: $P = [P_1;\cdots;P_g]$ using Matlab concatenation notation;
+assume the rows of $P$ are ordered to allow this.
+We can also assume that the rows of $Q$ and $W$ are ordered so that we can write $W = W_1 \directsum \cdots \directsum W_g$.
 

### 2009-07-24 paper.tex
...

@@
 The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
-Next, partition the rows of $P$ into small sets which can have simultaneously non-zero weights: $P = [P_1;\cdots;P_g]$ using Matlab concatenation notation;
-assume the rows of $P$ are ordered to allow this.
-We can also assume that the rows of $Q$ and $W$ are ordered so that we can write $W = W_1 \directsum \cdots \directsum W_g$.
+Moreover, we can partition the rows of $P$ into disjoint classes such that $w_{ij_1}$ and $w_{ij_2}$ are both non-zero only if $j_1$ and $j_2$ are in the same class.
+Thus, each row of $Q$ is associated with exactly one class, and all the points associated with a class lie in the subspace spanned by its rows in $P$.
 
-% Mixture modeling expresses probability densities as convex combinations of constituent probability distributions:
-% \begin{align}\eqn{mixture-model}
-%   q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
-% \end{align}
-% Here $q_i$ and $p_j$ are density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
-% \Equation{mixture-model} says that each of the distributions $q_i$ is a convex mixture of the
-
-% The question of why we see low-dimensional linear structure in the feature-frequency representation of network traffic is deeply connected to the nature of such linear representations themselves.
-% But the answer to the question also requires us to posit a hypothesis about the nature of network traffic.
+This model is an ideal, and observed behaviors have noise introduced by sampling, imperfect adherence to the model, and outliers.
+Yet the properties of the model explain the structures seen in \Figure{svd}.
+However, this is only one hypothesis that could fit the data.
+Like any hypothesis, it must be tested.
+The rest of this paper, aside from providing a useful, practical application, serves as a hypothesis test:
+we try to recover the matrices $W$ and $P$ from our imperfect observations of $Q$ and use the recovered model to predict flow behaviors.
+If the recovered model makes accurate predictions, this provides evidence that our hypothesis is correct.
 

### 2009-07-24 paper.tex
...

@@
 
-Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a stochastic model of flow behavior.
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical model of flow behavior.
 % A mixture model, to be specific.
@@
 
-This model is an ideal, and observed behaviors have noise introduced by sampling, imperfect adherence to the model, and outliers.
-Yet the properties of the model explain the structures seen in \Figure{svd}.
+The properties of the model explain the structures seen in \Figure{svd}.
+% This model is an ideal, and observed behaviors have noise introduced by sampling, imperfect adherence to the model, and outliers.
 However, this is only one hypothesis that could fit the data.
 Like any hypothesis, it must be tested.
-The rest of this paper, aside from providing a useful, practical application, serves as a hypothesis test:
-we try to recover the matrices $W$ and $P$ from our imperfect observations of $Q$ and use the recovered model to predict flow behaviors.
-If the recovered model makes accurate predictions, this provides evidence that our hypothesis is correct.
-
-\vfill
-\break
+The rest of this paper, aside from providing a practical application, serves as a hypothesis test:
+we try to recover the matrices $W$ and $P$ from our imperfect observations of $Q$ and use the recovered model to predict real flow behaviors.
+If the recovered model can make accurate predictions, this provides evidence that our model and hypothesis are largely accurate.
 

### 2009-07-24 paper.tex
...

@@
 \maketitle
+
+\begin{abstract}
+We have observed that when network traffic behaviors are embedded into vector spaces in certain ways, they tend to exhibit a great deal of low-rank linear structure.
+This observation leads us to the hypothesis that flow behaviors follow a specific type of finite mixture model.
+Aside from being of theoretical interest, this model has practical consequences:
+it allows us, among other things, to make detailed predictions about the probabilities of various future flow behaviors after observing only a handful of a flow's initial packets.
+This practical application serves a dual function.
+It provides a highly useful tool of network management, especially in wireless networks where such predictions can be leveraged to better allocate scarce bandwidth resources.
+But it also provides evidence that the hypothesized model not only explains the observed linear structure, but is actually an accurate model of real network traffic.
+\end{abstract}
 

### 2009-07-27 paper.tex
...

@@
 It provides a highly useful tool of network management, especially in wireless networks where such predictions can be leveraged to better allocate scarce bandwidth resources.
-But it also provides evidence that the hypothesized model not only explains the observed linear structure, but is actually an accurate model of real network traffic.
+However, it also provides evidence that the hypothesized model not only explains the observed linear structure, but is an accurate model of real network traffic.
 \end{abstract}

### 2009-07-27 paper.tex
...

@@
 The rest of this paper, aside from providing a practical application, serves as a hypothesis test:
-we try to recover the matrices $W$ and $P$ from our imperfect observations of $Q$ and use the recovered model to predict real flow behaviors.
-If the recovered model can make accurate predictions, this provides evidence that our model and hypothesis are largely accurate.
+we try to recover the matrices $W$ and $P$ from our noisy and imperfect observations of $Q$ and use the recovered model to predict real flow behaviors.
+If the recovered model can make accurate predictions, this provides evidence that our model and hypothesis are valid.
 
 \section{Methodology}\sec{methodology}
+
+Our methodology for recovering the hypothesized linear structure of network traffic from trace samples uses a variety of cutting-edge and recently developed data-mining techniques.
+To extract linear subspace structure, we use Ma~\emph{et~al.}'s 2007 algorithm based on lossy compression of data points~\cite{Ma07}.
 

### 2009-07-27 paper.tex
...

@@
 
-\section{Model \& Hypothesis}
+\section{Model \& Hypothesis}\sec{model-hypothesis}
 
@@
 
-Our methodology for recovering the hypothesized linear structure of network traffic from trace samples uses a variety of cutting-edge and recently developed data-mining techniques.
-To extract linear subspace structure, we use Ma~\emph{et~al.}'s 2007 algorithm based on lossy compression of data points~\cite{Ma07}.
+Our experimental methodology has three parts:
+training, prediction and evaluation.
+For training, we take a sample of network trace traffic and try to recover the structure hypothesized in \Section{model-hypothesis} from it.
+For prediction, we use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each.
+Finally, we must evaluate the quality of these predictions.
+As noted in the \Section{introduction}, due to nondeterminism we cannot possibly predict exact behaviors of flows and must instead evaluate how good a predicted distribution of possible behaviors is as compared to actual outcomes.
+% One relatively intuitive measure of prediction quality we use is threshold testing.
+
+% Our methodology for recovering the hypothesized linear structure of network traffic from trace samples uses a variety of powerful computational tools:
+% \begin{enumerate}
+%   \item Compute the linear representation matrix of the trace.
+%   \item Perform linear subspace segmentation, extracting subspaces near which significant amounts of data points lie.
+%   % For this step, we use Ma~\emph{et~al.}'s 2007 algorithm based on lossy compression of data~\cite{Ma07}.
+%   \item Use \caps{SVD} to determine an approximation rank for each segmentation and an optimal approximation basis.
+%   \item Compute the vertices of the intersection of each approximation subspace with the standard simplex.
+%   For each subspace, all nonnegative points can be written as convex linear combinations of these vertices.
+%   % For this we use Fukuda's excellent \caps{CDD}lib software for high-dimensional polyhedral computations, implementing a variety of convex hull and vertex enumeration algorithms~\cite{Avis92}.
+%   \item Find optimal weights to approximate each data point from the vertices of its subspace.
+% \end{enumerate}
 
@@
 \subsection{Flow Behavior Prediction}\sec{flow-behavior-prediction}
+
+\subsection{Evaluating Prediction Quality}\sec{evaluation}
 

### 2009-07-28 paper.tex
...
Add LaTeX for size/ival prediction CDFs.

@@
 
+\subsection{Data Sets}
+
 \subsection{Linear Representation}\sec{linear-representation}
+
+We model several aspects of flow behavior via feature-frequency linear representation.
+Packet sizes and and inter-packet intervals are represented using frequency histograms.
+Packet sizes are naturally quantized, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
 
@@
 
-\subsection{Evaluating Prediction Quality}\sec{evaluation}
+\subsection{Evaluating Predictions}\sec{evaluation}
 
 \section{Results}\sec{results}
+
+\begin{figure*}[t!]
+\begin{center}
+\subfloat[Packet sizes: \footnotesize{DARTMOUTH}]{%
+\includegraphics[width=3.545in]{size/dart}}
+\subfloat[Packet sizes: \footnotesize{SIGCOMM 2004}]{%
+\includegraphics[width=3.545in]{size/sc04}}
+\subfloat[Inter-packet intervals: \footnotesize{DARTMOUTH}]{%
+\includegraphics[width=3.545in]{ival/dart}}
+\subfloat[Inter-packet intervals: \footnotesize{SIGCOMM 2004}]{%
+\includegraphics[width=3.545in]{ival/sc04}}
+\caption{%
+CDFs of Kolmogorov-Smirnov p-values for various predictors of packet size distribution and inter-packet interval distribution for two selected traces.
+Lower curves are better and the ``ideal'' curve provides a lower bound on quality.
+Vertical segments indicate the farthest distance between the ideal and each CDF.
+This distance serves as a pessimistic measure of each predictor's error rate.
+}
+\fig{cdf}
+\end{center}
+\vspace{-2em}
+\end{figure*}
 

### 2009-07-28 paper.tex
...

@@
 
-\begin{figure*}[t!]
+\begin{figure*}[t]
 \begin{center}
@@
 \caption{%
-CDFs of Kolmogorov-Smirnov p-values for various predictors of packet size distribution and inter-packet interval distribution for two selected traces.
-Lower curves are better and the ``ideal'' curve provides a lower bound on quality.
+Empirical CDFs of Kolmogorov-Smirnov p-values for various predictors of packet size distribution and inter-packet interval distribution for two selected traces.
+Lower curves are better and the ideal curve provides a lower bound on quality.
 Vertical segments indicate the farthest distance between the ideal and each CDF.
@@
 
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=2.1in]{pkts_accuracy}
+\caption{Packet count prediction accuracy and uncertainty reduction.}
+\fig{packets-accuracy}
+\end{center}
+\vspace{-2em}
+\end{figure}
+
 \section{Discussion}\sec{discussion}

### 2009-07-28 paper.tex
...
Add datasets table.

@@
 \begin{figure*}[t]
+\vspace{-1em}
 \begin{center}
@@
 
+\begin{table*}
+\begin{center}
+\small
+\begin{tabular}{|c|c|c|c|c|c|c|}
+
+\hline
+\textbf{Trace} &
+\textbf{Year} &
+\textbf{Start} &
+\textbf{End} &
+\textbf{Type} &
+\textbf{Network} \\
+\hline
+
+{\footnotesize{DARTMOUTH}} &
+2003 &
+Mon Nov  3 18:39:48 &
+Wed Nov  5 16:09:48 &
+campus &
+Dartmouth College campus \\
+\hline
+
+{\footnotesize{IETF 60}} &
+2004 &
+Wed Aug  4 17:02:50 &
+Wed Aug  4 18:02:49 &
+conference &
+IETF hotel venue \\
+\hline
+
+{\footnotesize{IETF 67}} &
+2006 &
+Wed Nov  8 19:31:56 &
+Wed Nov  8 20:18:29 &
+conference &
+IETF hotel venue \\
+\hline
+
+{\footnotesize{SIGCOMM 2001}} &
+2001 &
+Wed Aug 29 17:24:47 &
+Fri Aug 31 19:53:15 &
+conference &
+SIGCOMM hotel venue \\
+\hline
+
+{\footnotesize{SIGCOMM 2004}} &
+2004 &
+Tue Aug 31 13:25:04 &
+Fri Sep  3 21:40:22 &
+conference &
+SIGCOMM hotel venue \\
+\hline
+
+{\footnotesize{UCSD}} &
+2007 &
+Thu Jan 11 07:59:50 &
+Fri Jan 12 07:59:55 &
+campus &
+UCSD engineering building \\
+\hline
+
+\end{tabular}
+\caption{Traffic traces used for analysis and experiments.}
+\tab{traces}
+\end{center}
+\vspace{-2em}
+\end{table*}
+
 \section{Methodology}\sec{methodology}
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \begin{center}
-\includegraphics[width=2.1in]{pkts_accuracy}
-\caption{Packet count prediction accuracy and uncertainty reduction.}
+\hfill
+\subfloat[Packet count: accuracy]{%
+\includegraphics[width=2.1in]{pkts_accuracy}}\hfill
+\subfloat[Packet count: uncertainty reduction]{%
+\includegraphics[width=2.1in]{pkts_certainty}}\hfill
+\subfloat[Packet count: prediction entropy]{%
+\includegraphics[width=2.1in]{pkts_entropy}}\hfill
+\caption{Packet count prediction accuracy, uncertainty reduction and entropy.}
 \fig{packets-accuracy}
@@
 \vspace{-2em}
-\end{figure}
+\end{figure*}
 

### 2009-07-28 paper.tex
Data sets and Feature-Frequency Representation.

@@
 
-\subsection{Linear Representation}\sec{linear-representation}
+We use randomly sampled traffic from six different network traces.
+All six traces are freely available from the \caps{CRAWDAD} wireless trace repository~\cite{Yeo06}.
+These traces represent a cross-section of traffic patterns over time and a reasonable variety of network types.
+Unfortunately, we could find no freely available commercial hotspot or corporate network traces that were usable for our analysis.
+We randomly sampled 5000 flows from each trace for training and another 5000 flows for testing.
 
-We model several aspects of flow behavior via feature-frequency linear representation.
-Packet sizes and and inter-packet intervals are represented using frequency histograms.
-Packet sizes are naturally quantized, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
+\subsection{Feature-Frequency Representation}\sec{representation}
+
+We embed the following features of flow behavior into vector spaces via feature-frequency linear representation:
+packet size distribution,
+inter-packet interval distribution,
+packet count,
+\caps{IP} protocol type,
+\caps{TCP/UDP} port numbers.
+All features are represented using frequency histograms.
+For packet size distribution, this is trivial since packet sizes are discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
+Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
+This scheme maps bins of interval durations to indices, with the bins starting at the scale of milliseconds, and ramping smoothly up to seconds at the high end of the scale.
+Intervals of more than ten minutes are considered to start a new flow.
+
+\caps{IP} protocol type is naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
+The ``histogram'' for a single flow is always just an indicator vector with a single unit entry.
+However, the linear embedding of this feature into a frequency space allows us to infer the distribution of possible values using linear analysis.
+Similarly, port numbers are naturally discrete.
+However, the range of port numbers (0-65535) is too large for practical representation using a full histogram.
+Instead, we represent port numbers by their remainder modulo 1000.
+Port numbers are sufficiently spread out that this still provides ample distinction between traffic types.
+We also represent the packet count of each flow.
+Since the potential packet count is unlimited, we must take some approach to limiting the dimension of this frequency representation.
+We use the simplest approach and simply cap the packet count at 3000 packets:
+all flows with 3000 or more packets fall into the same histogram bin.
+In practice, flows longer than 3000 packets are ``long.''
+
+Each flow is thus represented as a direct sum of frequency vectors:
+\begin{math}
+  \text{size} \directsum
+  \text{ival} \directsum
+  \text{type} \directsum
+  \text{ports} \directsum
+  \text{pkts} \in \R^{6256}.
+\end{math}
+A collection of flows is represented as a ``stacked'' matrix of row vectors:
+\begin{align}
+  X = [ ~
+  \text{Size} ~
+  \text{Ival} ~
+  \text{Type} ~
+  \text{Ports} ~
+  \text{Pkts} ~
+  ] \in \R^{m \times 6256}.
+\end{align}
 

### 2009-07-28 paper.tex
...

@@
 \begin{abstract}
-We have observed that when network traffic behaviors are embedded into vector spaces in certain ways, they tend to exhibit a great deal of low-rank linear structure.
-This observation leads us to the hypothesis that flow behaviors follow a specific type of finite mixture model.
-Aside from being of theoretical interest, this model has practical consequences:
+We have observed that when network traffic behaviors are represented in vector spaces as relative frequency histograms of behavioral features, they tend to exhibit a great deal of low-rank linear structure.
+We hypothesize that this structure is due the distribution of flow behaviors following a  finite mixture model.
+Aside from being of theoretical interest, this hypothesis has practical consequences:
 it allows us, among other things, to make detailed predictions about the probabilities of various future flow behaviors after observing only a handful of a flow's initial packets.
@@
 It provides a highly useful tool of network management, especially in wireless networks where such predictions can be leveraged to better allocate scarce bandwidth resources.
-However, it also provides evidence that the hypothesized model not only explains the observed linear structure, but is an accurate model of real network traffic.
+However, it also provides evidence that the hypothesized model provides a correct explanation for the observed linear structure in real network traffic.
 \end{abstract}
@@
 This paper is the first in a three-part series employing a numerical linear algebra techniques to understand and analyze network traffic patterns at a detailed level of individual flows and packets.\flownote
-All three papers employ the same theoretical framework but investigate different fundamental applications of traffic modeling:
+All three papers employ the same analytical framework but investigate different fundamental applications of traffic modeling:
 prediction, classification and generation.
@@
 
-The fundamental concept underlying this work is \emph{linear representation of network traffic}~\cite{Karpinski08}.
-A linear representation is a function that maps sets of flows to vectors, the coordinates of which describe aspects of flow behavior;
-it must distribute additively over set unions.
-% the representation must also satisfy a simple linearity condition which we will describe in detail later.
-% the sum of the representations of a set of flows is a vector that represents the aggregate behavior of the set of flows.
-An entire traffic trace may be represented as a matrix whose rows are the representations of flows in the trace.
-We restrict our consideration to a particularly natural and powerful class of representations whose coordinates are the frequencies of flow features in a set of flows.
-For example, the representation can have a dimension for each packet size up to the maximum transfer unit (\caps{MTU});
-coordinates in these dimensions indicate the count of packets of each size.
-Port numbers, \caps{IP} protocol type, and inter-packet intervals can be handled similarly.
-% Examples of flow features are the number of packets of a certain size or the use of a particular port number.
-% The fundamental premise of this work, and of the sequel, is that the representatives of actual flow behaviors can be understood as being sampled from
+This work begins with a particular way of representing flow behaviors as vectors.
+% and the behaviors of entire traffic traces as large, sparse matrices.
+% This approach was first proposed in by Karpinski~\emph{et~al.}~\cite{Karpinski08}, who called this general type of representation \emph{linear representations}.
+The representation is quite simple.
+For each feature of a flow, we represent that aspect of the flow's behavior as a \emph{feature-frequency vector}:
+a vector having a dimension for each possible value of the feature and whose coordinates are the relative frequency of values.
+For example, the vector representing the distribution of packet sizes of a flow having four 40-byte and two 145-bytes packets is
+\begin{align}
+  \frac{1}{6}\parens{4\vec{e}_{40} + 2\vec{e}_{145}}.
+\end{align}
+% The dimension of this representation is 1500 because that is typically the maximum transfer unit (\caps{MTU}) of local networks.
+Different aspects of flow behavior can be represented in this way, and those representations can be combined simply by taking the direct sum of their representation vectors.
 
-The essential observation of this research is that under such representations, real flows tend to lie near a small set of low-rank subspaces.
+The behavior of any set of flows, such as a traffic trace, can be represented as a matrix of feature-frequency vectors.
+We use the convection that features values correspond to matrix columns while flows correspond to matrix rows.
+When traffic traces are represented like this, a very curious thing happens:
+the resulting matrices exhibit a great deal of linear structure.
+Specifically, flow behaviors tend to lie near the union of a small set of low-rank subspaces.
+It is the investigation of this linear structure that lies at the core of this paper.
+
+We hypothesize that the reason for the linear structure is that the distribution of each flows behavior is a finite mixture of a small set of ``basic behaviors.''
+Moreover, the mixing of basic behaviors is itself structured.
 This property allows us to perform a drastic model reduction on trace traffic patterns while preserving the essential characteristics of the original traffic.
-Moreover, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:
+Furthermore, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:
 it allows us to use a least squares optimization on observed features to predict behavior.

### 2009-07-28 paper.tex
...

@@
 % The dimension of this representation is 1500 because that is typically the maximum transfer unit (\caps{MTU}) of local networks.
-Different aspects of flow behavior can be represented in this way, and those representations can be combined simply by taking the direct sum of their representation vectors.
+Different aspects of flow behavior can be represented in this way, and these representations can be combined by taking the direct sum of their representation vectors.
 
@@
 
-We hypothesize that the reason for the linear structure is that the distribution of each flows behavior is a finite mixture of a small set of ``basic behaviors.''
-Moreover, the mixing of basic behaviors is itself structured.
-This property allows us to perform a drastic model reduction on trace traffic patterns while preserving the essential characteristics of the original traffic.
+% We hypothesize that the reason for the linear structure is that the distribution of each flows behavior is a finite mixture of a small set of ``basic behaviors.''
+% Moreover, the mixing of basic behaviors is itself structured.
+The linear structure of network traffic allows us to perform a drastic model reduction on trace traffic patterns while preserving the essential characteristics of the original trace.
 Furthermore, when observable characteristics of flows are separated from features to be predicted, this model reduction is similar to a classical linear regression:
-it allows us to use a least squares optimization on observed features to predict behavior.
-We use this technique to predict the following characteristics of each flow from observing only a handful of its initial packets:
+it allows us to use least squares optimization on observed features to predict individual flow behavior.
+We use this technique to predict the following characteristics of flows from a few initial packets:
 \begin{enumerate}
@@
 
-At this point, it may be wise to address what we mean here by prediction.
-Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
-flow behavior is nondeterministic---flows may exhibit the same initial behavior but subsequently behave quite differently.
-The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
-How then can one measure the quality of a predicted distribution of outcomes?
-This subtle and difficult issue is central to our evaluation methodology.
+This prediction technique, is clearly of immediate practical use:
+applications range from network management to routing to quality of service.
+Looking deeper than these immediate applications, however, the mere fact that this prediction technique works at all provides strong evidence that the underlying hypothesized model of flow behavior is valid.
+Our future works will provide further applications and evidence by applying the same model to the problems of traffic classification and realistic workload generation.
+
+% At this point, it may be wise to address what we mean here by prediction.
+% Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
+% flow behavior is nondeterministic---flows may exhibit the same initial behavior but subsequently behave quite differently.
+% The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+% When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
+% How then can one measure the quality of a predicted distribution of outcomes?
+% This subtle and difficult issue is central to our evaluation methodology.
 

### 2009-07-28 paper.tex
...

@@
 
+The rest of this paper is organized as follows.
+In \Section{background}, we provide some background and the present the observations which motivate this work.
+Our model and hypothesis about flow behavior is presented in \Section{hypothesis}.
+The experimental methodology for testing this hypothesis is presented in \Section{methodology}, and the results follow in \Section{results}.
+\Section{discussion} discusses the meaning and impact of our results.
+\Section{related-work} gives context for this work in relation other research.
+We conclude with final remarks in \Section{conclusions}.
+
 % At this point, it may be wise to address what we mean here by prediction.
@@
 % This subtle and difficult issue is central to our evaluation methodology.
-
-% real-world network traces tend to have approximately low-rank representation matrices despite the potential rank of such matrices being very large.
-% Moreover, most flows' vectors lie on only one of a few even lower rank subspaces.
-% Moreover, the low-rank subspace near which real traffic lies can be decomposed into subspaces of even lower rank, with almost all flows lying in one of these subspaces.
-% Typically, these smaller subspaces are one-dimensional, but in some cases they may have up to a dozen dimensions.
-
-% the vast majority of observed flow behaviors in any given trace lie in one of a handful low-rank subspaces.
-% These subspaces are mostly of rank one, but a few have dimensions as high as ten.
-
-% Karpinski~\emph{et~al.}~\cite{Karpinski08} further observed that the simplifications commonly used in traffic modeling correspond to multiplication of traffic representation matrices by low-rank matrices, averaging over rows or columns.
-% They demonstrated experimentally that such simplifications severely distort important behavioral properties of traffic traces;
-% viewing these simplification as low-rank matrix transformations, this is unsurprising:
-% most of the detail of the original traffic behavior is destroyed.
-% It was suggested that matrix factorization provides an alternative and superior means of model reduction.
 
@@
 
-\section{Model \& Hypothesis}\sec{model-hypothesis}
+\section{Model \& Hypothesis}\sec{model}\sec{hypothesis}
 
@@
 training, prediction and evaluation.
-For training, we take a sample of network trace traffic and try to recover the structure hypothesized in \Section{model-hypothesis} from it.
+For training, we take a sample of network trace traffic and try to recover the structure hypothesized in \Section{hypothesis} from it.
 For prediction, we use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each.
@@
 
+\section{Related Work}\sec{related-work}
+
 \section{Conclusions}\sec{conclusions}

### 2009-07-28 paper.tex
...

@@
 
-This prediction technique, is clearly of immediate practical use:
+This prediction technique is clearly of immediate practical use:
 applications range from network management to routing to quality of service.
-Looking deeper than these immediate applications, however, the mere fact that this prediction technique works at all provides strong evidence that the underlying hypothesized model of flow behavior is valid.
-Our future works will provide further applications and evidence by applying the same model to the problems of traffic classification and realistic workload generation.
+Looking deeper than these applications, however, the mere fact that this prediction technique works at all provides evidence that the underlying hypothesized model of flow behavior is valid.
+Our future works will provide evidence and applications by applying this model to the problems of traffic classification and realistic workload generation.
 
@@
 Our model and hypothesis about flow behavior is presented in \Section{hypothesis}.
-The experimental methodology for testing this hypothesis is presented in \Section{methodology}, and the results follow in \Section{results}.
-\Section{discussion} discusses the meaning and impact of our results.
-\Section{related-work} gives context for this work in relation other research.
+The experimental methodology for testing this hypothesis is presented in \Section{methodology} while the results follow in \Section{results}.
+\Section{discussion} discusses the meaning and impact of our results and
+\Section{related-work} gives context for this work in relation prior research.
 We conclude with final remarks in \Section{conclusions}.

### 2009-07-28 paper.tex
...

@@
 \begin{align}
-  \frac{1}{6}\parens{4\vec{e}_{40} + 2\vec{e}_{145}}.
+  \text{size} = \frac{1}{6}\parens{4\vec{e}_{40} + 2\vec{e}_{145}}.
 \end{align}
@@
 
-A linear representation, as proposed by Karpinski~\emph{et~al.}~\cite{Karpinski08}, maps sets of flows into a vector space such that disjoint union of sets becomes addition in the vector space:
+\newfootnote{\directsumnote}{The direct sum is the vector space analogue of a cross product of sets.}
+
+Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors. spaces.
+The low-level details of this representation are given in \Section{representation}.
+For now, what the reader must know is that each feature is represented by a vector where each possible value of the feature correspond to a dimension and the normalized frequency of that value occurring is the corresponding coordinate in the vector.
+We have already given the example of packet size distribution in the introduction.
+The vectors are normalized in that the sum of the coordinates of each feature-frequency vector is unity.
+To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
 \begin{align}
-  \Phi(s_1 \disunion s_2) = \Phi(s_1) + \Phi(s_2).
+  \text{flow} =
+  \text{size} \directsum
+  \text{ival} \directsum
+  \text{type} \directsum
+  \text{ports} \directsum
+  \text{pkts}.
 \end{align}
-The representation function should map each set of flows to a vector that in some sense describes their collective behavior.
-For example: let $\Phi(s) \in \R^1$ be the number of packets in the flows in $s$.
-Since the number of packets in the disjoint union of sets of flows is the sum of the number in each set individually, this representation is linear.
-Given two representations, their direct sum is readily seen to also be a linear representation:
+The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
 \begin{align}
-  \Phi(s) = \Phi_1(s) \directsum \Phi_2(s).
+  \text{Size} = \begin{bmatrix}
+    \text{size}_1 \\
+    \vdots \\
+    \text{size}_m
+  \end{bmatrix}.
 \end{align}
-
-A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
-A 1500-dimensional vector can be used to count the frequency of packet sizes seen, from one byte up to the typical \caps{MTU} of 1500 bytes.
-In this representation, $\Phi(s)$ has as its $i$th entry the number of packets in $s$ of size $i$.
-Since \caps{IP} protocols range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
-Histogram representations for port numbers and inter-packet intervals are detailed in \Section{methodology}.
-% Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
-% For practical purposes, however, this is too large a vector space, so we represent port numbers in 1000-dimensions, and reduce port numbers modulo 1000.
-% Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
-The representation we use is a direct sum of these histogram representations.
-% The combination is a representation in 5256 dimensions.
-An entire trace is represented as a matrix of such representation vectors---one flow per row.
-This paper explores the linear structure of such matrices.
+The overall behavior of the collection of flows then becomes a concatenation of these feature matrices:
+\begin{align}
+  X = [ ~
+  \text{Size} ~
+  \text{Ival} ~
+  \text{Type} ~
+  \text{Ports} ~
+  \text{Pkts} ~
+  ].
+\end{align}
+It is this matrix, representing the total behavior of a collection of flows, to which we apply our analysis.
 
@@
 
-To see the linear structure of these representation matrices, we apply one of the most powerful tools of linear algebra:
+To see the linear structure of these representation matrices, we apply one of the most fundamental tools of linear algebra:
 the singular value decomposition (\caps{SVD}).
@@
 Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
-This property allows us to visualize the structure of $X$ using initial columns of $US$.
-However, we are interested only interested in direction since
-vectors differing only by magnitude lie in the same one-dimensional subspace.
-To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane from the origin.\projectionnote
+This property allows us to visualize the structure of $X$ using initial columns of $US$---and this is precisely what we do.
+However, it must be done in a particular way.
+First determine an appropriate value for $k$.
+Here we take the number of dimensions necessary to account for 75\% of the variance seen in $X$;
+other approaches work well too.
+Discarding the rest of the dimensions has the effect of ``denoising'' the data.
+The vectors of the remaining $k$-dimensional reduction of $X$ may have different magnitudes, but we are only interested in direction, since vectors differing only be magnitude lie in the same one-dimensional subspace.
+Therefore, we project each row vector through the origin onto the unit-sum hyperplane.\projectionnote
+It is the coordinates of this reduced, projected data which exhibit clear linear structure.
 
-\Figure{svd} illustrates the results of applying \caps{SVD} dimension reduction to sample matrices of six different traffic traces.
-The figures show scatter plots of the most significant two projected \caps{SVD} coordinates for a matrix of 5000 sample flows from each trace.
-% The original data encodes packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows from each trace.
-% The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
-Linear structure is visible in all six:
+\Figure{svd} shows scatter plots of most significant two coordinates when this procedure is applied to a random sample of 5000 flows from each of six unrelated traffic traces.
+Clear linear structure is visible in all six traces:
 flow behaviors are heavily concentrated near low-dimensional subspaces.
@@
 behaviors on higher-dimensional subspaces appear as ``smears.''
-% There is a smattering of outliers, not especially near such structures.
 
@@
 Neither question can be answered fully in a single paper, but we will begin to address the existential problem in the next section.
-In the rest of the paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
+In the rest of this paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
 

### 2009-07-28 paper.tex
...

@@
 
-Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical model of flow behavior.
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical statistical model for flow behavior.
 % A mixture model, to be specific.
@@
 
-The properties of the model explain the structures seen in \Figure{svd}.
+The properties of the model exactly explain the structures seen in \Figure{svd}.
 % This model is an ideal, and observed behaviors have noise introduced by sampling, imperfect adherence to the model, and outliers.
-However, this is only one hypothesis that could fit the data.
+However, this is only one possible hypothesis that fits the data.
 Like any hypothesis, it must be tested.

### 2009-07-28 paper.tex
...

@@
 
-% At this point, it may be wise to address what we mean here by prediction.
-% Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
-% flow behavior is nondeterministic---flows may exhibit the same initial behavior but subsequently behave quite differently.
-% The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-% When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
-% How then can one measure the quality of a predicted distribution of outcomes?
-% This subtle and difficult issue is central to our evaluation methodology.
-
 \section{Background \& Motivation}\sec{background}\sec{motivation}
@@
 
-\Figure{svd} shows scatter plots of most significant two coordinates when this procedure is applied to a random sample of 5000 flows from each of six unrelated traffic traces.
+\Figure{svd} shows scatter plots of most significant two coordinates when this procedure is applied to a random sample of 5000 flows from each of six independent traffic traces.
 Clear linear structure is visible in all six traces:
@@
 
-The properties of the model exactly explain the structures seen in \Figure{svd}.
-% This model is an ideal, and observed behaviors have noise introduced by sampling, imperfect adherence to the model, and outliers.
+This model explains the structures found in \Figure{svd}.
+Points lying along the same structure are in the same class, and each structure is ``generated'' by a small set of vertices:
+the points belonging to a structure are near the convex hull generated by those vertices.
 However, this is only one possible hypothesis that fits the data.
@@
 training, prediction and evaluation.
-For training, we take a sample of network trace traffic and try to recover the structure hypothesized in \Section{hypothesis} from it.
-For prediction, we use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each.
-Finally, we must evaluate the quality of these predictions.
-As noted in the \Section{introduction}, due to nondeterminism we cannot possibly predict exact behaviors of flows and must instead evaluate how good a predicted distribution of possible behaviors is as compared to actual outcomes.
-% One relatively intuitive measure of prediction quality we use is threshold testing.
+For training, we take a sample of network trace traffic and try to recover the model hypothesized in \Section{hypothesis} from it.
+Next, we use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each flow.
+Finally, we must evaluate the quality of these predictions by comparing them to the actual behavior of those flows.
 
-% Our methodology for recovering the hypothesized linear structure of network traffic from trace samples uses a variety of powerful computational tools:
-% \begin{enumerate}
-%   \item Compute the linear representation matrix of the trace.
-%   \item Perform linear subspace segmentation, extracting subspaces near which significant amounts of data points lie.
-%   % For this step, we use Ma~\emph{et~al.}'s 2007 algorithm based on lossy compression of data~\cite{Ma07}.
-%   \item Use \caps{SVD} to determine an approximation rank for each segmentation and an optimal approximation basis.
-%   \item Compute the vertices of the intersection of each approximation subspace with the standard simplex.
-%   For each subspace, all nonnegative points can be written as convex linear combinations of these vertices.
-%   % For this we use Fukuda's excellent \caps{CDD}lib software for high-dimensional polyhedral computations, implementing a variety of convex hull and vertex enumeration algorithms~\cite{Avis92}.
-%   \item Find optimal weights to approximate each data point from the vertices of its subspace.
-% \end{enumerate}
+It seems wise at this point to address the nature of prediction for highly nondeterministic phenomena like network traffic.
+Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
+flows may exhibit the same initial behavior but subsequently behave quite differently.
+The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
+How then can one measure the quality of a predicted distribution of outcomes?
+This subtle and difficult issue is central to our evaluation methodology.
 
@@
 
-We embed the following features of flow behavior into vector spaces via feature-frequency linear representation:
+We embed the following features of flow behavior into vector spaces via feature-frequency representation:
 packet size distribution,
 inter-packet interval distribution,
-packet count,
 \caps{IP} protocol type,
-\caps{TCP/UDP} port numbers.
-All features are represented using frequency histograms.
-For packet size distribution, this is trivial since packet sizes are discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
+\caps{TCP/UDP} port numbers,
+and packet count.
+Packet sizes are already naturally discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
 Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
@@
 Intervals of more than ten minutes are considered to start a new flow.
-
-\caps{IP} protocol type is naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
-The ``histogram'' for a single flow is always just an indicator vector with a single unit entry.
-However, the linear embedding of this feature into a frequency space allows us to infer the distribution of possible values using linear analysis.
-Similarly, port numbers are naturally discrete.
-However, the range of port numbers (0-65535) is too large for practical representation using a full histogram.
+\caps{IP} protocol type also naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
+The feature-frequency vector for a single flow in this case is always just a vector with a single unit entry and all other entries zero.
+Port numbers are naturally discrete as well, but the range of port numbers (0-65535) is too large for practical representation using a full histogram.
 Instead, we represent port numbers by their remainder modulo 1000.
 Port numbers are sufficiently spread out that this still provides ample distinction between traffic types.
+Moreover, since the direction of a flow is generally uninteresting, we represent both source and destination in a single vector containing two values.
 We also represent the packet count of each flow.
-Since the potential packet count is unlimited, we must take some approach to limiting the dimension of this frequency representation.
+Since the potential packet count is unlimited, some approach is needed for limiting the dimension of this frequency representation.
 We use the simplest approach and simply cap the packet count at 3000 packets:
-all flows with 3000 or more packets fall into the same histogram bin.
-In practice, flows longer than 3000 packets are ``long.''
-
-Each flow is thus represented as a direct sum of frequency vectors:
-\begin{math}
-  \text{size} \directsum
-  \text{ival} \directsum
-  \text{type} \directsum
-  \text{ports} \directsum
-  \text{pkts} \in \R^{6256}.
-\end{math}
-A collection of flows is represented as a ``stacked'' matrix of row vectors:
-\begin{align}
-  X = [ ~
-  \text{Size} ~
-  \text{Ival} ~
-  \text{Type} ~
-  \text{Ports} ~
-  \text{Pkts} ~
-  ] \in \R^{m \times 6256}.
-\end{align}
+all flows with 3000 or more packets fall into the same histogram bin and are simply considered ``very large.''
 

### 2009-07-29 paper.tex
...

@@
 
-Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical statistical model for flow behavior.
-% A mixture model, to be specific.
-We propose viewing each flow as having a probability distribution of possible behaviors it could exhibit.
-% We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
-We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
-Moreover, only even smaller subsets of these basic behaviors are typically combined with each other.
-Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
-\begin{align}\eqn{mixture-model}
-  q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
-\end{align}
-Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
-\Equation{mixture-model} is expressed succinctly as matrix multiplication.
-Writing $Q_{ik} = q_i(k)$, $W_{ij} = w_{ij}$, and $P_{jk} = p_j(k)$, we have:
-\begin{align}\eqn{mixture-model-matrix}
-  Q = WP.
-\end{align}
-The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
-Moreover, we can partition the rows of $P$ into disjoint classes such that $w_{ij_1}$ and $w_{ij_2}$ are both non-zero only if $j_1$ and $j_2$ are in the same class.
-Thus, each row of $Q$ is associated with exactly one class, and all the points associated with a class lie in the subspace spanned by its rows in $P$.
-
-This model explains the structures found in \Figure{svd}.
-Points lying along the same structure are in the same class, and each structure is ``generated'' by a small set of vertices:
-the points belonging to a structure are near the convex hull generated by those vertices.
-However, this is only one possible hypothesis that fits the data.
-Like any hypothesis, it must be tested.
-The rest of this paper, aside from providing a practical application, serves as a hypothesis test:
-we try to recover the matrices $W$ and $P$ from our noisy and imperfect observations of $Q$ and use the recovered model to predict real flow behaviors.
-If the recovered model can make accurate predictions, this provides evidence that our model and hypothesis are valid.
-
 \begin{table*}
@@
 
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical statistical model for flow behavior.
+% A mixture model, to be specific.
+We propose viewing each flow as having a probability distribution of possible behaviors it could exhibit.
+% We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
+We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
+Moreover, only even smaller subsets of these basic behaviors are typically combined with each other.
+Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
+\begin{align}\eqn{mixture-model}
+  q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
+\end{align}
+Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
+\Equation{mixture-model} is expressed succinctly as matrix multiplication.
+Writing $Q_{ik} = q_i(k)$, $W_{ij} = w_{ij}$, and $P_{jk} = p_j(k)$, we have:
+\begin{align}\eqn{mixture-model-matrix}
+  Q = WP.
+\end{align}
+The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
+Moreover, we can partition the rows of $P$ into disjoint classes such that $w_{ij_1}$ and $w_{ij_2}$ are both non-zero only if $j_1$ and $j_2$ are in the same class.
+Thus, each row of $Q$ is associated with exactly one class, and all the points associated with a class lie in the subspace spanned by its rows in $P$.
+
+This model explains the structures found in \Figure{svd}.
+Points lying along the same structure are in the same class, and each structure is ``generated'' by a small set of vertices:
+the points belonging to a structure are near the convex hull generated by those vertices.
+However, this is only one possible hypothesis that fits the data.
+Like any hypothesis, it must be tested.
+The rest of this paper, aside from providing a practical application, serves as a hypothesis test:
+we try to recover the matrices $W$ and $P$ from our noisy and imperfect observations of $Q$ and use the recovered model to predict real flow behaviors.
+If the recovered model can make accurate predictions, this provides evidence that our model and hypothesis are valid.
+
 \section{Methodology}\sec{methodology}
@@
 All six traces are freely available from the \caps{CRAWDAD} wireless trace repository~\cite{Yeo06}.
-These traces represent a cross-section of traffic patterns over time and a reasonable variety of network types.
-Unfortunately, we could find no freely available commercial hotspot or corporate network traces that were usable for our analysis.
+Details of the traces are shown in \Table{traces}.
+These traces represent a broad cross-section of traffic patterns over time and a reasonable variety of network types.
+Unfortunately, we could not find and freely available commercial hotspot or corporate network traces that were usable for our analysis.
 We randomly sampled 5000 flows from each trace for training and another 5000 flows for testing.
@@
 \subsection{Subspace Segmentation}\sec{subspace-segmentation}
+
+Once the data have been expressed in matrix form, the next step is attempt to detect the linear structures seen in \Figure{svd}.
+% Recall that these correspond to the point classes in our hypothesized model.
+We use Ma~\emph{et~al.}'s~\cite{Ma07} algorithm for segmenting multivariate data into a union of subspaces using lossy data coding and compression.
+The essential idea of this algorithm is to compute the number of bits needed to encode the data under various assumed groupings and their implied distributions.
+The grouping which leads to the least overall encode size is the desired segmentation.
+This algorithm has many desirable qualities:
+unlike most clustering algorithms, it finds subspaces, not full-rank regions;
+it is highly robust to outliers and noisy data.
+This algorithm has one parameter:
+the maximum allowable distortion of the data, $\varepsilon^2$.
+We follow Ma's lead when applying their algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
+The result of this stage is a segmentation of the data into classes of rows, precisely as our hypothesized model requires.
 

### 2009-07-29 paper.tex
...

@@
   \text{type} \directsum
-  \text{ports} \directsum
+  \text{port} \directsum
   \text{pkts}.
@@
   \text{Type} ~
-  \text{Ports} ~
+  \text{Port} ~
   \text{Pkts} ~
@@
 Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
-This scheme maps bins of interval durations to indices, with the bins starting at the scale of milliseconds, and ramping smoothly up to seconds at the high end of the scale.
+This scheme maps bins of interval durations to indices, with the bins starting at the scale of milliseconds, but ramping smoothly up to seconds at the high end of the scale.
 Intervals of more than ten minutes are considered to start a new flow.
@@
 \subsection{Subspace Factorization}\sec{subspace-factorization}
+
+\newfootnote{\svdnnnote}{The first singular vector of a nonnegative matrix is guaranteed to be nonnegative, but the next will generally not be.}
+
+The next step towards recovering the structure of the hypothesized model is to determine the hull points of each linear substructure.
+We have determined, as well as we can, which of the $q_i$ belong to the same structures.
+Now we must determine the distributions $p_j$ which generate these structures.
+We can use \caps{SVD} to determine the most prominent linear components of each structure, but \caps{SVD} will give us vectors of mixed sign.\svdnnnote
+many invalid points will lie within the convex hull of these vectors.
+\caps{SVD} can give orthogonal vectors spanning the structure, but we must turn to other techniques to recover possible vectors for the $p_j$.
+
+In essence, what we have is a nonnegative matrix factorization (\caps{NMF}) problem:
+if $Q_*$ is a sub-matrix of rows in the same structure class, we want to find nonnegative matrices, $W_*$ and $P_*$, such that $Q_* \approx W_* P_*$;
+% We have the additional constraint that $W_*$ and $P_*$ have unit-sum rows, but that can easily be guaranteed since 
+$P$ is then a vertical concatenation of small $P_*$ matrices, while $W$ is a direct sum of $W_*$ matrices.
+Many \caps{NMF} algorithms have been proposed since Lee and Seung published the first~\cite{Lee01}.
+We use Kim and Park's alternating non-negative least squares (\caps{ANLS}) algorithm~\cite{Kim08}.
+It converges much more rapidly than the Lee and Seung algorithm, but stalls when it gets close to an optimum, so we refine that result by applying the Lee and Seung algorithm.
+
+\newfootnote{\cddnote}{To do this, we use Avis and Fukuda's vertex enumeration algorithm as implemented in Fukuda's excellent CDD library~\cite{Avis92}.}
+
+We find that using standard random \caps{NMF} initialization does not yield factorizations of sufficient quality, so we have developed the following initialization procedure.
+Reduce the data to the number structural dimensions sought using \caps{SVD}.
+Cluster the reduced data points into one cluster per structural dimension using $k$-means clustering.
+Use cluster membership from this step to combine the original data, leaving only one composite row per desired structural dimension.
+These rows provide a ``denoised'' set of nonnegative spanning vectors that should recover the true low-dimensional subspace from which the data were sampled.
+Finally, we find the intersection of this subspace with the standard simplex in $\R^n$.
+This step gives us a set of vertices geometrically guaranteed to contain all of the recovered low-dimensional subspace.\cddnote
 

### 2009-07-29 paper.tex
...

--- 2009-07-29 paper.tex
@@
 
-\subsection{Subspace Segmentation}\sec{subspace-segmentation}
+\subsection{Model Recovery}\sec{model-recovery}
+
+\subsubsection{Subspace Segmentation}\sec{subspace-segmentation}
 
@@
 We use Ma~\emph{et~al.}'s~\cite{Ma07} algorithm for segmenting multivariate data into a union of subspaces using lossy data coding and compression.
-The essential idea of this algorithm is to compute the number of bits needed to encode the data under various assumed groupings and their implied distributions.
-The grouping which leads to the least overall encode size is the desired segmentation.
-This algorithm has many desirable qualities:
-unlike most clustering algorithms, it finds subspaces, not full-rank regions;
-it is highly robust to outliers and noisy data.
-This algorithm has one parameter:
+The essential idea of this algorithm is to compute the number of bits needed to encode the data under various assumed groupings.
+The grouping which yields the best compression is the desired segmentation.
+This algorithm has many desirable qualities.
+Two are essential:
+1) unlike most clustering algorithms, it finds subspaces, not regions;
+2) it is highly robust to outliers and noisy data.
+The algorithm has one parameter:
 the maximum allowable distortion of the data, $\varepsilon^2$.
-We follow Ma's lead when applying their algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
+We follow Ma's lead when applying the algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
 The result of this stage is a segmentation of the data into classes of rows, precisely as our hypothesized model requires.
 
-\subsection{Subspace Factorization}\sec{subspace-factorization}
+\subsubsection{Subspace Factorization}\sec{subspace-factorization}
+
+\newfootnote{\svdnnnote}{The first singular vector of a nonnegative matrix is guaranteed to be nonnegative, but the next will generally not be.}
+
+The next step towards recovering the structure of the hypothesized model is to determine the hull points of each linear substructure.
+We have determined, as well as we can, which $q_i$ belong to the same structures.
+Now we must determine the vertices, $p_j$, which generate these structures.
+We can use \caps{SVD} to determine the most prominent linear components of each structure, but \caps{SVD} will give us vectors of mixed sign.\svdnnnote
+Since negative probability distributions are meaningless, these do not give us a plausible recovery of our model.
+We must turn to other techniques to recover possible vectors for the $p_j$.
+
+In essence, what we have is a nonnegative matrix factorization (\caps{NMF}) problem:
+if $Q_*$ is a sub-matrix of rows in the same structure class, we want to find nonnegative matrices, $W_*$ and $P_*$, such that $Q_* \approx W_* P_*$.~
+% We have the additional constraint that $W_*$ and $P_*$ have unit-sum rows, but that can easily be guaranteed since 
+$P$ is a vertical concatenation of the $P_*$ matrices, while $W$ is a direct sum of $W_*$ matrices.
+Many \caps{NMF} algorithms have been proposed since Lee and Seung published the first~\cite{Lee01}.
+We use Kim and Park's alternating non-negative least squares (\caps{ANLS}) algorithm~\cite{Kim08} first for its rapid initial convergence, but refine the result using Lee and Seung's Euclidean algorithm.
+
+\newfootnote{\cddnote}{To do this, we use Avis and Fukuda's vertex enumeration algorithm as implemented in Fukuda's excellent CDD library~\cite{Avis92}.}
+
+We find that using the standard random \caps{NMF} initialization does not yield factorizations of sufficient quality, so we have developed the following initialization procedure.
+Reduce the data to the number structural dimensions sought using \caps{SVD}.
+Cluster the reduced data points into one cluster per structural dimension using $k$-means clustering.
+Use cluster membership from this step to combine the original data, leaving only one composite row per desired structural dimension.
+These rows provide a ``denoised'' set of nonnegative spanning vectors that estimate the true low-dimensional subspace from which the data were sampled.
+Finally, we find the intersection of this subspace with the standard simplex in $\R^n$.\,\cddnote
+This step gives us a set of nonnegative vertices geometrically guaranteed to contain all of the estimated low-dimensional subspace.
+These vectors are used to initialize the \caps{NMF} algorithms, yielding extremely good factorization for small substructures.
 
 \subsection{Flow Behavior Prediction}\sec{flow-behavior-prediction}
+
+In order to perform flow behavior prediction.
 

### 2009-07-29 paper.tex
wip

@@
 To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
-\begin{align}
+\begin{align}\eqn{flow}
   \text{flow} =
@@
 The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
-\begin{align}
+\begin{align}\eqn{Size}
   \text{Size} = \begin{bmatrix}
@@
 The overall behavior of the collection of flows then becomes a concatenation of these feature matrices:
-\begin{align}
-  X = [ ~
+\begin{align}\eqn{X}
+  X = \bracket{ ~
     \text{Size} ~
@@
     \text{Pkts} ~
-  ].
+  }.
 \end{align}
@@
 Many \caps{NMF} algorithms have been proposed since Lee and Seung published the first~\cite{Lee01}.
-We use Kim and Park's alternating non-negative least squares (\caps{ANLS}) algorithm~\cite{Kim08} first for its rapid initial convergence, but refine the result using Lee and Seung's Euclidean algorithm.
+We use Kim and Park's alternating non-negative least squares (\caps{ANLS}) algorithm~\cite{Kim08:anls} first for its rapid initial convergence, but refine the result using Lee and Seung's Euclidean algorithm.
 
@@
 
-In order to perform flow behavior prediction.
+In order to perform flow behavior prediction, we separate the flow features into the observable ones and the ones that must be predicted.
+For example, if we want to predict total flow behavior from some number of initial packets, then we would have the following matrix of observables:
+\begin{align}
+  X_\text{o} = \bracket{ ~
+    \text{Size}_{\text{init}} ~
+    \text{Ival}_{\text{init}} ~
+    \text{Type} ~
+    \text{Port} ~
+  }
+\end{align}
+The matrix of ``predictables'' would be:
+\begin{align}
+  X_\text{p} = \bracket{ ~
+    \text{Size}_{\text{rest}} ~
+    \text{Ival}_{\text{rest}} ~
+    \text{Pkts} ~
+  }
+\end{align}
+We use the packet sizes and inter-packet intervals of the initial packets, together with their \caps{IP} type and port numbers to predict the distribution of packets sizes and inter-packet intervals for the rest of the flow, as well as predicting a distribution of possible of packet counts.
+In order to do this prediction, subspace factorization must be done with the observable and predictable matrices separated.
+That is, subspace factorization must be done on the matrix $[X_\text{o}~X_\text{p}]$.
+% However, we perform our subspace segmentation step on the combined data as in \Equation{X}.
+% Why the difference? We have two pragmatic reasons:
+% 1) the segmentation step seems to work better with the unseparated matrix.
+% 2) we want to apply our prediction technique with different numbers of initial packets;
+% the segmentation step is very slow, while the factorization step is relatively fast;
+% this way we can use the same segmentation result for varying numbers of initial packets.
+
+Our technique is fundamentally similar to fitting a regression line to noisy two-dimensional training data and then using the $x$-coordinates of test data points to linearly predict the $y$-coordinates.
+The main differences are the vastly higher number of dimensions and more complex structure we assume for our data.
+From the training data, we recover our estimated version of the $P$ matrix.
+We split this into its observable and predictable parts: $P_\text{o}$ and $P_\text{p}$.
+Given an observation matrix, $X^*_\text{o}$, to perform prediction, we find the nonnegative matrix of weights, $W_*$, minimizing the squared Frobenius error:
+\begin{align}
+  \norm{X^*_\text{o} - W^* P_\text{o}}^2_\text{frob}.
+\end{align}
+This is equivalent to minimizing the square error for each observation individually, but optimizing \emph{en masse} allows fast non-negative least squares algorithms to be used~\cite{Benthem04,Kim08:block-pivot}.
+The predicted behavior for these flows is simply weight matrix times the predictable feature matrix:
+\begin{align}
+  X^*_\text{p} = W^* P_\text{p}.
+\end{align}
+The rows of $X^*_\text{p}$ are our prediction about the behavior of the flows observed in $X^*_\text{o}$.
+Each row contains a packet size distribution, an inter-packet interval distribution and a distribution of packet counts.
+
+% With our hypothetical model parameters, $p_j$, extracted from training data, for each test observation, $q^*_\text{o}$, we find nonnegative weights, $w^*_j$, that minimize the squared error
+% \begin{align}\textstyle
+%   \norm{q^*_\text{o}-\sum {w^*_j p_j}}^2.
+% \end{align}
+% The predicted behavior of the flow is simply
 

### 2009-07-29 paper.tex
...

@@
 
-Our experimental methodology has three parts:
+Our experimental procedure has three parts:
 training, prediction and evaluation.
-For training, we take a sample of network trace traffic and try to recover the model hypothesized in \Section{hypothesis} from it.
-Next, we use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each flow.
+For training, we take a sample of network trace traffic and attempt to recover the model hypothesized in \Section{hypothesis} from it.
+We use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each flow.
 Finally, we must evaluate the quality of these predictions by comparing them to the actual behavior of those flows.
+In this section we present the methodology for training and prediction;
+how to evaluate prediction results against actual behavior is left until the next section together with the results.
 
-It seems wise at this point to address the nature of prediction for highly nondeterministic phenomena like network traffic.
-Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
-flows may exhibit the same initial behavior but subsequently behave quite differently.
-The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
-How then can one measure the quality of a predicted distribution of outcomes?
-This subtle and difficult issue is central to our evaluation methodology.
+% It seems wise at this point to address the nature of prediction for highly nondeterministic phenomena like network traffic.
+% Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
+% flows may exhibit the same initial behavior but subsequently behave quite differently.
+% The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+% When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
+% How then can one measure the quality of a predicted distribution of outcomes?
+% This subtle and difficult issue is central to our evaluation methodology.
 

### 2009-07-29 paper.tex
...

@@
 These traces represent a broad cross-section of traffic patterns over time and a reasonable variety of network types.
-Unfortunately, we could not find and freely available commercial hotspot or corporate network traces that were usable for our analysis.
+Unfortunately, we could not find any freely available residential or corporate network traces that were usable for our analysis.
 We randomly sampled 5000 flows from each trace for training and another 5000 flows for testing.
@@
 The algorithm has one parameter:
-the maximum allowable distortion of the data, $\varepsilon^2$.
+the maximum allowable distortion of the data
 We follow Ma's lead when applying the algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
@@
 
-\newfootnote{\svdnnnote}{The first singular vector of a nonnegative matrix is guaranteed to be nonnegative, but the next will generally not be.}
+\newfootnote{\svdnnnote}{The first singular vector of a nonnegative matrix is guaranteed to be nonnegative, but the other singular vectors will have mixed sign.}
 
@@
 This is equivalent to minimizing the square error for each observation individually, but optimizing \emph{en masse} allows fast non-negative least squares algorithms to be used~\cite{Benthem04,Kim08:block-pivot}.
-The predicted behavior for these flows is simply weight matrix times the predictable feature matrix:
+The predicted behavior for these flows is simply the weight matrix times the predictable feature matrix:
 \begin{align}
@@
 \end{align}
-The rows of $X^*_\text{p}$ are our prediction about the behavior of the flows observed in $X^*_\text{o}$.
+The rows of $X^*_\text{p}$ are our predictions about the behavior of flows observed in $X^*_\text{o}$.
 Each row contains a packet size distribution, an inter-packet interval distribution and a distribution of packet counts.
@@
 
-\subsection{Evaluating Predictions}\sec{evaluation}
-
 \section{Results}\sec{results}
@@
 \end{center}
-\vspace{-2em}
+\vspace{-1em}
 \end{figure*}

### 2009-07-29 paper.tex
...

@@
 
-% It seems wise at this point to address the nature of prediction for highly nondeterministic phenomena like network traffic.
-% Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
-% flows may exhibit the same initial behavior but subsequently behave quite differently.
-% The best we can do then is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-% When compared with a specific outcome, such a prediction cannot be said to be ``wrong'' or ``right.''
-% How then can one measure the quality of a predicted distribution of outcomes?
-% This subtle and difficult issue is central to our evaluation methodology.
-
 \subsection{Data Sets}
@@
 
+% It seems wise at this point to address the nature of prediction for highly nondeterministic phenomena like network traffic.
+Brief reflection shows that one cannot hope to accurately predict the exact behavior of flows:
+flows may exhibit the same initial behavior but subsequently behave quite differently.
+The best we can hope to do is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
+When compared with a specific outcome, such a prediction cannot be said to be unequivocally ``wrong'' or ``right.''
+How then can one measure the quality of a predicted distribution of outcomes?
+This subtle and difficult issue is central to our evaluation methodology.
+
 \begin{figure*}[t]

### 2009-07-29 paper.tex
wip

@@
 These rows provide a ``denoised'' set of nonnegative spanning vectors that estimate the true low-dimensional subspace from which the data were sampled.
-Finally, we find the intersection of this subspace with the standard simplex in $\R^n$.\,\cddnote
+Finally, we find the intersection of this subspace with the standard simplex in $\R^n$.\cddnote
 This step gives us a set of nonnegative vertices geometrically guaranteed to contain all of the estimated low-dimensional subspace.
@@
 The best we can hope to do is to predict a \emph{distribution} of possible outcomes given the observed initial behavior.
-When compared with a specific outcome, such a prediction cannot be said to be unequivocally ``wrong'' or ``right.''
-How then can one measure the quality of a predicted distribution of outcomes?
+When compared with a specific outcome, such a prediction cannot be said to be unequivocally right or wrong.
+How then can we measure the quality of a predicted distribution of outcomes as compared to a single outcome?
 This subtle and difficult issue is central to our evaluation methodology.
+
+% To compare the predicted distributions of packet sizes and inter-packet intervals against the observed distributions, we use the most common ``goodness-of-fit'' test:
+
+To evaluate the quality of predicted distributions as compared with a sample that may or may not come from that distribution, we use the single-sample Kolmogorov-Smirnov (K-S) ``goodness-of-fit'' test~\cite{Feller68}.
+The K-S test compares the cumulative distribution function (\caps{CDF}) of a theoretical distribution with the empirical cumulative distribution function of a sample of values, such as packet sizes.
+The test uses the largest absolute difference between the two \caps{CDF}s as a statistic to measure how plausible it is that the sample was drawn from the theoretical distribution.
+The K-S statistic can be transformed into a p-value:
+the probability of obtaining a result at least as extreme as the observed result, assuming that the sample did come from the theoretical distribution.
+
+We evaluate the quality of our packet size and inter-packet interval distribution predictions by looking at the distribution of K-S p-values for the predicted distribution versus the observed sample.
+In theory, if the predictions are all correct, the p-values should be uniformly distributed on the interval [0,1].
+In practice, however, the K-S test is not perfect, and its p-values are not uniform.
+Instead of comparing predictions to the theoretical ideal of uniformity, we use a more pragmatic ideal as a baseline:
+p-values of samples that are actually drawn from the predicted distributions.
+We measure quality of predictions by how much worse that this ideal the predictions are.
 

### 2009-07-29 paper.tex
wip

@@
 
+\newfootnote{\ksidealnote}{If the K-S test were ideal, this pragmatic baseline would in fact be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
+
 We evaluate the quality of our packet size and inter-packet interval distribution predictions by looking at the distribution of K-S p-values for the predicted distribution versus the observed sample.
@@
 p-values of samples that are actually drawn from the predicted distributions.
-We measure quality of predictions by how much worse that this ideal the predictions are.
+We measure quality of predictions by how much worse that this ideal the predictions are.\ksidealnote
+
+\Figure{cdf} shows the the \caps{CDF}s p-values for our ideal baseline and the predicted packet size and inter-packet interval distributions.
+For comparison, we also show the p-values results for two simplistic prediction techniques.
+The first alternative prediction technique is to always predicting the mean behavior of the entire training trace.
+This does not perform as badly as one might think it would---always guessing the average is not an abysmal failure.
+The second alternative technique is to just guess what you've already seen.
+We call this ``self-prediction.''
+This is likely the best guess one can make about a flow, without having learned from any other flows:
+just guess that it will keep doing what it has already done.
 

### 2009-07-29 paper.tex
...

--- 2009-07-29 paper.tex
@@
 Vertical segments indicate the farthest distance between the ideal and each CDF.
-This distance serves as a pessimistic measure of each predictor's error rate.
+This distance serves as a measure of each predictor's overall error rate.
 }
@@
 \end{center}
-\vspace{-1em}
+\vspace{-1.5em}
 \end{figure*}
@@
 To evaluate the quality of predicted distributions as compared with a sample that may or may not come from that distribution, we use the single-sample Kolmogorov-Smirnov (K-S) ``goodness-of-fit'' test~\cite{Feller68}.
-The K-S test compares the cumulative distribution function (\caps{CDF}) of a theoretical distribution with the empirical cumulative distribution function of a sample of values, such as packet sizes.
+The K-S test compares the cumulative distribution function (\caps{CDF}) of a theoretical distribution with the empirical cumulative distribution function of a sample of values.
 The test uses the largest absolute difference between the two \caps{CDF}s as a statistic to measure how plausible it is that the sample was drawn from the theoretical distribution.
 The K-S statistic can be transformed into a p-value:
-the probability of obtaining a result at least as extreme as the observed result, assuming that the sample did come from the theoretical distribution.
+the probability of obtaining a result at least as extreme as the observed result, assuming that the sample really did come from the theoretical distribution.
 
-We evaluate the quality of our packet size and inter-packet interval distribution predictions by looking at the distribution of K-S p-values for the predicted distribution versus the observed sample.
+\newfootnote{\ksidealnote}{If the K-S test were ideal, this pragmatic baseline would in fact be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
+
+We evaluate the quality of our packet size and inter-packet interval distribution predictions by exampining the distribution of K-S p-values for the predicted distribution versus the observed sample.
 In theory, if the predictions are all correct, the p-values should be uniformly distributed on the interval [0,1].
 In practice, however, the K-S test is not perfect, and its p-values are not uniform.
-Instead of comparing predictions to the theoretical ideal of uniformity, we use a more pragmatic ideal as a baseline:
+Instead of comparing predictions to the nonexistent theoretical ideal, we use a more pragmatic ideal as a baseline:
 p-values of samples that are actually drawn from the predicted distributions.
-We measure quality of predictions by how much worse that this ideal the predictions are.
+We measure quality of predictions by how much worse that this ideal the predictions are.\ksidealnote
+
+\Figure{cdf} shows selected p-value \caps{CDF}s for our predicted distributions compared to the ideal baseline.
+For comparison, we also show the results for two simplistic prediction techniques.
+The first alternative---``mean''---always predicts the average behavior of the entire training trace.
+This does not perform as badly as one might think it would:
+always guessing the mean is not an abysmal failure.
+The second alternative---``self''---predicts the behavior that has already been seen.
+% This is the best guess one can make about a flow, without having been trained on any other flows:
+% just guess that it will keep doing what it has already done.
+
+A na\"ive point-measure of the error of a prediction technique is the rate at which it passes the K-S test at a predetermined significance level, such as 1\% or 5\%, as compred to the ideal.
+This is equivalent to measuring the distance above the ideal \caps{CDF} at the 0.01 or 0.05 p-value mark.
+This measure of error is na\"ive because these traditional significance levels are intentionally highly conservative:
+they are chosen so as to incorrectly reject the null only 1\% or 5\% of the time.
+Thus they only reject very bad predictions.
+The mean predictor makes a perfect example:
+by guessing the average behavior it actually makes relatively few really bad predictions but makes far more poor to mediocre predictions than it should, and almost no good predictions.
+Our predictor, on the other hand, maintains a fairly steady error rate somewhat above the ideal.
+
+As better point-measure of the quality of predictors, we propose the maximum vertical distance between its p-value \caps{CDF} and the ideal.
+This is equivalent to choosing the worst possible predetermined significance level for each predictor and evaluating it at that level.
+Under this measure, predictors cannot ``hide'' their errors by shifting them around the p-value range.
+The maximum vertical difference from the ideal for each predictor is shown in \Figure{cdf} with vertical bars.
+The number next to each bar is the length of this distance:
+this is the the overall error rate of the predictor.
+
+\begin{table*}
+\begin{center}
+\small
+\begin{tabular}{|c|c|c|c|c|c|c|}
+
+\hline
+\textbf{Trace} &
+\textbf{Predictor} &
+\textbf{0.01-error} &
+\textbf{0.05-error} &
+\textbf{max-error} \\
+\hline
+
+\end{tabular}
+\caption{Error rates for size and interval predictions for all traces.}
+\tab{traces}
+\end{center}
+\vspace{-2em}
+\end{table*}
 

### 2009-07-29 paper.tex
Simplified prediction stats table.
Added plot for success rates.

@@
 The maximum vertical difference from the ideal for each predictor is shown in \Figure{cdf} with vertical bars.
-The number next to each bar is the length of this distance:
-this is the the overall error rate of the predictor.
+This is the the overall error rate of the predictor.
+The success rate is the complement of the error rate.
+\Figure{success-rates} shows the success rates of various predictions across the six data sets.
 
-\begin{table*}
+\begin{figure}[!b]
+\vspace{-2em}
 \begin{center}
-\small
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\includegraphics[width=3.5in]{pred_stats}
+\caption{Success rates of various prediction methods.}
+\fig{success-rates}
+\end{center}
+\vspace{-1em}
+\end{figure}
 
-\hline
-\textbf{Trace} &
-\textbf{Predictor} &
-\textbf{0.01-error} &
-\textbf{0.05-error} &
-\textbf{max-error} \\
-\hline
-
-\end{tabular}
-\caption{Error rates for size and interval predictions for all traces.}
-\tab{traces}
-\end{center}
-\vspace{-2em}
-\end{table*}
+% \begin{table}[b]
+% \begin{center}
+% \small
+% \begin{tabular}{|c|c|c|c|c|c|c|}
+% \hline
+% &
+% \multicolumn{2}{c|}{\textbf{Predicted}} &
+% \multicolumn{2}{c|}{\textbf{Self}} &
+% \multicolumn{2}{c|}{\textbf{Mean}} \\
+% \hline
+% \textbf{Trace} &
+% \textbf{size} & \textbf{ival} &
+% \textbf{size} & \textbf{ival} &
+% \textbf{size} & \textbf{ival} \\
+% \hline
+% \input{pred_stats}
+% \hline
+% \end{tabular}
+% \caption{Error rates for size and interval predictions.}
+% \tab{error-rates}
+% \end{center}
+% \vspace{-1em}
+% \end{table}
 

### 2009-07-29 paper.tex
...

@@
 
-As better point-measure of the quality of predictors, we propose the maximum vertical distance between its p-value \caps{CDF} and the ideal.
+As a better point-measure of the quality of predictors, we propose the maximum vertical distance between its p-value \caps{CDF} and the ideal.
 This is equivalent to choosing the worst possible predetermined significance level for each predictor and evaluating it at that level.
@@
 \begin{figure}[!b]
-\vspace{-2em}
+\vspace{-1em}
 \begin{center}
@@
 \end{center}
-\vspace{-1em}
+\vspace{-0.9em}
 \end{figure}
@@
 \begin{figure*}[t]
+\vspace{-1em}
 \begin{center}
-\hfill
 \subfloat[Packet count: accuracy]{%
-\includegraphics[width=2.1in]{pkts_accuracy}}\hfill
+\includegraphics[width=2.1in]{pkts_accuracy}}
 \subfloat[Packet count: uncertainty reduction]{%
-\includegraphics[width=2.1in]{pkts_certainty}}\hfill
+\includegraphics[width=2.1in]{pkts_certainty}}
 \subfloat[Packet count: prediction entropy]{%
-\includegraphics[width=2.1in]{pkts_entropy}}\hfill
-\caption{Packet count prediction accuracy, uncertainty reduction and entropy.}
+\includegraphics[width=2.1in]{pkts_entropy}}
+\caption{Packet count predictions: accuracy, uncertainty reduction, and entropy.}
 \fig{packets-accuracy}
@@
 \end{figure*}
+
+Next we turn to the evaluation of packet counts predictions.
+Here we cannot use the K-S test:
+although we have a predicted distribution of possible packet counts, we have only one value to compare it to---the actual number of packets.
+The K-S test is useless for samples of fewer than three values---it cannot invalidate any outcome.
+Instead we use the predicted distributions themselves to yield p-values for each packet count.
+We compute how ``extreme'' each observed packet count value is according to the predicted distribution of packet count values:
+that is, what fraction of the distribution is as large or larger than the observed value.
+If the predictions are accurate, then these ``p-values'' should be uniformly distributed.
+
+Accuracy alone is insufficient, however:
+guessing the mean distribution of packet counts yields a perfectly accurate predictor according to the same criterion.
+What we want, however, is to remain accurate but become more certain about the packet count.
 

### 2009-07-29 paper.tex
...

@@
 \begin{center}
-\subfloat[Packet count: accuracy]{%
+\subfloat[Packet count: accuracy]{\fig{packets-accuracy}%
 \includegraphics[width=2.1in]{pkts_accuracy}}
-\subfloat[Packet count: uncertainty reduction]{%
+\subfloat[Packet count: uncertainty]{\fig{packets-uncertainty}%
 \includegraphics[width=2.1in]{pkts_certainty}}
-\subfloat[Packet count: prediction entropy]{%
+\subfloat[Packet count: entropy]{\fig{packets-entropy}%
 \includegraphics[width=2.1in]{pkts_entropy}}
-\caption{Packet count predictions: accuracy, uncertainty reduction, and entropy.}
-\fig{packets-accuracy}
+\caption{Packet count predictions: accuracy, uncertainty reduction, and distribution entropy.}
 \end{center}
@@
 Accuracy alone is insufficient, however:
-guessing the mean distribution of packet counts yields a perfectly accurate predictor according to the same criterion.
+guessing the mean distribution of packet counts yields a perfectly accurate predictor according to this criterion.
 What we want, however, is to remain accurate but become more certain about the packet count.
+We can measure this certainty by computing the information-theoretic entropy of each prediction.
+We want our accuracy to remain intact, but the entropy of our predictions to go down as much as possible.
+A related measure is the reduction in uncertainty:
+the exponential of reduction in entropy from the mean predictor.
+This is the fraction of the original ``number'' of possible values that remain after our prediction as compared to before.
+We put ``number'' in quotes because even though entropy is measured in bits, it doesn't count enumerable discrete values like a byte value would.
+
+\Figure{packets-accuracy} shows the \caps{CDF} of p-values of observed packet counts versus our predicted distributions.
+While some of the \caps{CDF}s do not remain perfectly straight along the diagonal, overall the tendency is for our predictions to remain accurate.
+There is some tendency for the observations to fall at the high and low ends of our predictions, indicating that we are in some cases making predictions that are too narrow.
+Still the general adherence to the diagonal shows that oh the whole, the packet predictions remain fairly accurate.
+\Figure{packets-uncertainty} shows that while we are not losing much accuracy, we are gaining a great deal of certainty about what the actual packet count will be.
+Most of the time, 50\% of the time, we lose all but 20\% of our original uncertainty.
+\Figure{packets-entropy} complements this figure, showing what the actual distribution of entropy values for the predictors are.
 

### 2009-07-29 paper.tex
...

@@
 \Figure{packets-uncertainty} shows that while we are not losing much accuracy, we are gaining a great deal of certainty about what the actual packet count will be.
-Most of the time, 50\% of the time, we lose all but 20\% of our original uncertainty.
-\Figure{packets-entropy} complements this figure, showing what the actual distribution of entropy values for the predictors are.
+Most of the time, 50\% of the time, we discard around 80\% of the original uncertainty.
+\Figure{packets-entropy} complements this figure, showing what the actual distribution of entropy values for the six data sets.
 

### 2009-07-29 paper.tex
...

@@
 Aside from being of theoretical interest, this hypothesis has practical consequences:
-it allows us, among other things, to make detailed predictions about the probabilities of various future flow behaviors after observing only a handful of a flow's initial packets.
+it allows us, among other things, to make detailed, accurate predictions about the probabilities of various future flow behaviors after observing only a handful of a flow's initial packets.
+% We are able to predict the distribution of future packet sizes and inter-packet intervals from only five initial packets with between 70\% and 90\% accuracy across a variety of network traces.
+% For half of flows, we can reduce the uncertainty of our packet count predictions by 80\% or more from the same observations.
 This practical application serves a dual function.
@@
 This subtle and difficult issue is central to our evaluation methodology.
+% In what follows, we have performed prediction from observing only five initial packets from each flow.
 
@@
 
+Our results show that we can predict the packet size distribution and inter-packet interval distribution of flows from only observing a handful of initial packets of each flow.
+Specifically, we have used five initial packets to predict these distributions with between 70\% and 90\% accuracy across a wide variety of network traces.
+Moreover, we can reduce our uncertainty about the number of packets that flows will have by 80\% or more, about half of the time.
+Immediate applications of this ability to accurately predict individual flow behaviors abound.
+Quality of service (\caps{QoS}) schemes attempting to provide high-quality service guarantees for streaming voice and video wireless through wireless networks could use such predictions to know---almost presciently---how much data flows will transmit and how often.
+
+The real importance of this work has little to do with these potential practical applications.
+The ability to predict flow behavior is merely a pleasant side effect of the real breakthrough:
+a detailed statistical model for network-wide flow behaviors.
+The fundamental concept underlying this work is that there are a relatively small collection of ``basic behaviors'' that generate the vast majority of the flow behaviors that we actually see in any particular trace.
+These behaviors are associated with distributions of various flow features:
+packet sizes, inter-packet intervals, \caps{IP} protocol, port numbers and packet counts.
+these features are drawn from these distributions to ``create'' actual flows.
+
+% TODO: mention the two dimension structure assumption.
+% TODO: mention the five packets in the results somewhere.
+
 \section{Related Work}\sec{related-work}

### 2009-07-29 paper.tex
...

@@
 packet sizes, inter-packet intervals, \caps{IP} protocol, port numbers and packet counts.
-these features are drawn from these distributions to ``create'' actual flows.
+these features are drawn from these distributions to ``construct'' actual flows.
+We have treated the features as though they are sampled independently, but no doubt they are not really independent at all.
+
+Part of the reason the model of flow generation as mixtures of a basic set of behaviors is not obvious and has not previously been noted is likely that with convex linear combinations, so many different combinations can be created.
+Moreover, the analytical tools to extract this structure have only existed for a few years.
+In our model recovery procedure, we use a subspace segmentation algorithm published in 2007~\cite{Ma07}, two \caps{NMF} algorithms published in 2001 and 2008, respectively~\cite{Lee01,Kim08:anls}, and a low-rank \caps{NMF} initialization technique published for the first time here.
+And these techniques are still quite immature:
+they will only improve over time, yielding even better recovery of the model parameters.
+Such improvements will immediately improve the performance of the prediction methods employed in this paper.
+
+It is our hope and belief that this model of flow behavior will provide future insights into other aspects of network traffic modeling:
+in particular traffic classification and realistic workload generation.
+The recovered model used in this paper is very close to providing a usable model for traffic generation already.
+The largest missing piece is a model for realistic weight matrices.
+If such applications based on this flow behavior model or other similar ones prove as fruitful as the application of flow behavior prediction, that will provide further evidence of the effectiveness and validity of our proposed model.
 

### 2009-07-30 paper.tex
Tweaking various plots.
Work on the abstract.

@@
 \begin{abstract}
-We have observed that when network traffic behaviors are represented in vector spaces as relative frequency histograms of behavioral features, they tend to exhibit a great deal of low-rank linear structure.
+We have observed that when network traffic behaviors are represented in vector spaces as relative frequency histograms of behavioral features, they exhibit low-rank linear structure.
 We hypothesize that this structure is due the distribution of flow behaviors following a  finite mixture model.
@@
 it allows us, among other things, to make detailed, accurate predictions about the probabilities of various future flow behaviors after observing only a handful of a flow's initial packets.
-% We are able to predict the distribution of future packet sizes and inter-packet intervals from only five initial packets with between 70\% and 90\% accuracy across a variety of network traces.
-% For half of flows, we can reduce the uncertainty of our packet count predictions by 80\% or more from the same observations.
-This practical application serves a dual function.
-It provides a highly useful tool of network management, especially in wireless networks where such predictions can be leveraged to better allocate scarce bandwidth resources.
-However, it also provides evidence that the hypothesized model provides a correct explanation for the observed linear structure in real network traffic.
+We are able to predict the distribution of future packet sizes and inter-packet intervals from only five initial packets with between 70\% and 90\% accuracy across a variety of network traces;
+for about half of flows, we can reduce uncertainty about packet count by 80\% or more.
+These practical applications serve dual functions.
+They provide highly useful tools for network management, routing decisions, and quality of service schemes.
+However, they also provide evidence that the hypothesized model gives a correct explanation for the observed linear structure in real network traffic.
 \end{abstract}
@@
 \subfloat[Packet sizes: \footnotesize{DARTMOUTH}]{%
-\includegraphics[width=3.545in]{size/dart}}
+\includegraphics[width=2.35in]{size/dart}}
+\subfloat[Packet sizes: \footnotesize{IETF 67}]{%
+\includegraphics[width=2.35in]{size/ie67}}
 \subfloat[Packet sizes: \footnotesize{SIGCOMM 2004}]{%
-\includegraphics[width=3.545in]{size/sc04}}
+\includegraphics[width=2.35in]{size/sc04}}
 \subfloat[Inter-packet intervals: \footnotesize{DARTMOUTH}]{%
-\includegraphics[width=3.545in]{ival/dart}}
+\includegraphics[width=2.35in]{ival/dart}}
+\subfloat[Inter-packet intervals: \footnotesize{IETF 67}]{%
+\includegraphics[width=2.35in]{ival/ie67}}
 \subfloat[Inter-packet intervals: \footnotesize{SIGCOMM 2004}]{%
-\includegraphics[width=3.545in]{ival/sc04}}
+\includegraphics[width=2.35in]{ival/sc04}}
 \caption{%
@@
 \subfloat[Packet count: accuracy]{\fig{packets-accuracy}%
-\includegraphics[width=2.1in]{pkts_accuracy}}
+\includegraphics[width=2.351in]{pkts_accuracy}}
 \subfloat[Packet count: uncertainty]{\fig{packets-uncertainty}%
-\includegraphics[width=2.1in]{pkts_certainty}}
+\includegraphics[width=2.351in]{pkts_certainty}}
 \subfloat[Packet count: entropy]{\fig{packets-entropy}%
-\includegraphics[width=2.1in]{pkts_entropy}}
+\includegraphics[width=2.351in]{pkts_entropy}}
 \caption{Packet count predictions: accuracy, uncertainty reduction, and distribution entropy.}

### 2009-07-31 paper.tex
...

@@
 This paper is the first in a three-part series employing a numerical linear algebra techniques to understand and analyze network traffic patterns at a detailed level of individual flows and packets.\flownote
-All three papers employ the same analytical framework but investigate different fundamental applications of traffic modeling:
+All three papers employ the same analytical approach but investigate different fundamental applications of traffic modeling:
 prediction, classification and generation.
-In this paper we first present the framework and then demonstrate how it can predict the behavior of individual flows from the observation of only a handful of their initial packets.
+In this paper we present our hypothesis and demonstrate how it can be used to predict the behavior of individual flows from observing only a handful of initial packets.
 % The following papers will focus on classification of flows according to application type and realistic workload generation.
@@
 The experimental methodology for testing this hypothesis is presented in \Section{methodology} while the results follow in \Section{results}.
-\Section{discussion} discusses the meaning and impact of our results and
-\Section{related-work} gives context for this work in relation prior research.
+\Section{discussion} discusses the meaning and impact of these results and
+\Section{related-work} gives context for this research in relation prior work.
 We conclude with final remarks in \Section{conclusions}.
@@
 
-Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors. spaces.
+Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors.
 The low-level details of this representation are given in \Section{representation}.
-For now, what the reader must know is that each feature is represented by a vector where each possible value of the feature correspond to a dimension and the normalized frequency of that value occurring is the corresponding coordinate in the vector.
+For now, what the reader must know is that each feature is represented by a vector where each dimension corresponds to a possible value of the feature;
+the coordinates of the vector are the normalized frequencies of the corresponding feature values.
 We have already given the example of packet size distribution in the introduction.
-The vectors are normalized in that the sum of the coordinates of each feature-frequency vector is unity.
-To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
+Vectors are normalized so the coordinates of each feature-frequency vector have unit sum.
+To combine multiple feature-frequency vectors into a composite description of the overall behavior of a flow, we take the direct sum individual feature representations:\directsumnote
 \begin{align}\eqn{flow}
@@
 \end{align}
-The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
+The behavior of a feature across a collection of flows can be expressed as a matrix where each row represents a flow:
 \begin{align}\eqn{Size}
@@
 
-\newfootnote{\svdnote}{More precisely, the product of the first $k$ columns of $U$ with the first $k$ rows of $\trans{V}$ is a Frobenius-optimal rank-$k$ approximation of $X$.}
+\newfootnote{\svdnote}{More precisely, the product of the first $k$ columns of $US$ with the first $k$ rows of $\trans{V}$ is a Frobenius-optimal rank-$k$ approximation of $X$.}
 
@@
 First determine an appropriate value for $k$.
-Here we take the number of dimensions necessary to account for 75\% of the variance seen in $X$;
+We use the number of dimensions necessary to account for 75\% of the variance seen in $X$;
 other approaches work well too.
 Discarding the rest of the dimensions has the effect of ``denoising'' the data.
-The vectors of the remaining $k$-dimensional reduction of $X$ may have different magnitudes, but we are only interested in direction, since vectors differing only be magnitude lie in the same one-dimensional subspace.
+The vectors of the remaining $k$-dimensional reduction of $X$ may have different magnitudes, but we are only interested in direction, since vectors differing only by magnitude lie in the same one-dimensional subspace.
 Therefore, we project each row vector through the origin onto the unit-sum hyperplane.\projectionnote
-It is the coordinates of this reduced, projected data which exhibit clear linear structure.
+It is the coordinates of this reduced, projected data which exhibit linear structure.
 
@@
 Followed, of course, by the practical question: \emph{How can we exploit this structure?}
-Neither question can be answered fully in a single paper, but we will begin to address the existential problem in the next section.
+Neither question can be answered fully in a single paper, but we will attempt to address the existential problem in the next section.
 In the rest of this paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
@@
 
-Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a hypothetical statistical model for flow behavior.
+Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as the hypothesis that flow behavior follows a specific statistical model.
 % A mixture model, to be specific.
@@
 The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
-Moreover, we can partition the rows of $P$ into disjoint classes such that $w_{ij_1}$ and $w_{ij_2}$ are both non-zero only if $j_1$ and $j_2$ are in the same class.
-Thus, each row of $Q$ is associated with exactly one class, and all the points associated with a class lie in the subspace spanned by its rows in $P$.
+Moreover, we can partition the rows of $P$ into classes such that $w_{ij_1}$ and $w_{ij_2}$ are both non-zero only if $j_1$ and $j_2$ are in the same class.
+Thus, each row of $Q$ is associated with exactly one class, and all the points associated with a class lie in the subspace spanned by its associated rows in $P$.
+Of course, even if this model is true, actual observed behaviors will be sampled from the ideal distributions of $Q$, producing an observation matrix, $X$, with a great deal of noise.
+% In particular, for features with vary small samples, the amount of noise introduced is quite large.
+Moreover, we expect that there should be many outliers:
+network traffic is such a complex phenomenon that no single model will explain all observations.
 
-This model explains the structures found in \Figure{svd}.
-Points lying along the same structure are in the same class, and each structure is ``generated'' by a small set of vertices:
-the points belonging to a structure are near the convex hull generated by those vertices.
+Despite these caveats, this model explains the structures found in \Figure{svd}.
+Points along the same low-rank structure are in the same class.
+Each structure is ``generated'' by a small set of vertices:
+the points belonging to the structure are near the hull of its vertices.
 However, this is only one possible hypothesis that fits the data.
@@
 training, prediction and evaluation.
-For training, we take a sample of network trace traffic and attempt to recover the model hypothesized in \Section{hypothesis} from it.
-We use the trained model to predict behaviors of a separate set of flows from the same trace, using only knowledge of a few initial packets of each flow.
-Finally, we must evaluate the quality of these predictions by comparing them to the actual behavior of those flows.
+For training, we take samples of network traces and attempt to recover the parameters of the model hypothesized in \Section{hypothesis}.
+We use the recovered model to predict behaviors of a separate set of flows from the same trace, using only knowledge of five initial packets from each flow.
+Finally, we must evaluate the quality of these predictions by comparing them against the actual behavior of those flows.
 In this section we present the methodology for training and prediction;
-how to evaluate prediction results against actual behavior is left until the next section together with the results.
+how to evaluate prediction results is described together with the results in \Section{results}.
 
@@
 We randomly sampled 5000 flows from each trace for training and another 5000 flows for testing.
+Since the initial five packets are used for prediction, we consider only flows with at least five packets.
 
@@
 and packet count.
-Packet sizes are already naturally discrete, taking on only values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
+%
+Packet sizes are naturally discrete, taking on values from 1 to the \caps{MTU}, which was 1500 in all of our traces.
+%
 Inter-packet intervals must be quantized, using the scheme described by Karpinski~\emph{et~al.}~\cite{Karpinski08} but using only 500 dimensions instead of 1500.
@@
 Intervals of more than ten minutes are considered to start a new flow.
-\caps{IP} protocol type also naturally discrete, with values from 0 to 255, represented by a 256-dimensions histogram vector.
-The feature-frequency vector for a single flow in this case is always just a vector with a single unit entry and all other entries zero.
+%
+\caps{IP} protocol type is also naturally discrete, with values from 0 to 255, represented by a 256-dimensional vector.
+The feature-frequency vector for a single flow is always just a vector with a single unit entry and zeros elsewhere.
+%
 Port numbers are naturally discrete as well, but the range of port numbers (0-65535) is too large for practical representation using a full histogram.
@@
 Moreover, since the direction of a flow is generally uninteresting, we represent both source and destination in a single vector containing two values.
+%
 We also represent the packet count of each flow.

### 2009-07-31 paper.tex
...

@@
 
-Once the data have been expressed in matrix form, the next step is attempt to detect the linear structures seen in \Figure{svd}.
+Once the data have been expressed in matrix form, the next step is try to detect the linear structures seen in \Figure{svd}.
 % Recall that these correspond to the point classes in our hypothesized model.
-We use Ma~\emph{et~al.}'s~\cite{Ma07} algorithm for segmenting multivariate data into a union of subspaces using lossy data coding and compression.
-The essential idea of this algorithm is to compute the number of bits needed to encode the data under various assumed groupings.
+We use Ma~\emph{et~al.}'s algorithm for segmenting multivariate data into a union of subspaces using lossy data coding and compression~\cite{Ma07}.
+The essential idea of this algorithm is to compute the number of bits needed to encode the data under various groupings.
 The grouping which yields the best compression is the desired segmentation.
-This algorithm has many desirable qualities.
-Two are essential:
+This algorithm has several desirable qualities.
+Two are essential for us:
 1) unlike most clustering algorithms, it finds subspaces, not regions;
@@
 The algorithm has one parameter:
-the maximum allowable distortion of the data
-We follow Ma's lead when applying the algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
+the maximum allowable distortion of the data.
+We follow the authors when applying the algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
 The result of this stage is a segmentation of the data into classes of rows, precisely as our hypothesized model requires.

### 2009-07-31 paper.tex
...

@@
 We follow the authors when applying the algorithm to \caps{SVD} data and use $\varepsilon^2 = 1$.
-The result of this stage is a segmentation of the data into classes of rows, precisely as our hypothesized model requires.
+The result of this stage is a segmentation of the $q_i$ into classes associated with structures, precisely as our hypothesized model requires.
 
@@
 
-The next step towards recovering the structure of the hypothesized model is to determine the hull points of each linear substructure.
-We have determined, as well as we can, which $q_i$ belong to the same structures.
+The next step towards recovering the structure of the hypothesized model is to determine the hull points of each linear structure.
+In the previous step, we have determined, as well as we can, which $q_i$ belong to the same structures.
 Now we must determine the vertices, $p_j$, which generate these structures.
@@
 In essence, what we have is a nonnegative matrix factorization (\caps{NMF}) problem:
-if $Q_*$ is a sub-matrix of rows in the same structure class, we want to find nonnegative matrices, $W_*$ and $P_*$, such that $Q_* \approx W_* P_*$.~
+if $Q_*$ is a sub-matrix of rows in the same structure class, we want to find nonnegative matrices, $W_*$ and $P_*$, such that $Q_* \approx W_* P_*$.
 % We have the additional constraint that $W_*$ and $P_*$ have unit-sum rows, but that can easily be guaranteed since 
-$P$ is a vertical concatenation of the $P_*$ matrices, while $W$ is a direct sum of $W_*$ matrices.
+Large $P$ is a vertical concatenation of these $P_*$ matrices, while $W$ is row-permutation of the direct sum of $W_*$ matrices.
 Many \caps{NMF} algorithms have been proposed since Lee and Seung published the first~\cite{Lee01}.
@@
 This step gives us a set of nonnegative vertices geometrically guaranteed to contain all of the estimated low-dimensional subspace.
-These vectors are used to initialize the \caps{NMF} algorithms, yielding extremely good factorization for small substructures.
+These vectors are used to initialize the \caps{NMF} algorithms, yielding extremely good factorization for small structures.
 

### 2009-07-31 paper.tex
...

@@
 
-In order to perform flow behavior prediction, we separate the flow features into the observable ones and the ones that must be predicted.
+To predict flow behavior, we must separate flow features into those that are observed and those that must be predicted.
 For example, if we want to predict total flow behavior from some number of initial packets, then we would have the following matrix of observables:
@@
   X_\text{o} = \bracket{ ~
-    \text{Size}_{\text{init}} ~
-    \text{Ival}_{\text{init}} ~
+    \text{Size}_\text{init} ~
+    \text{Ival}_\text{init} ~
     \text{Type} ~
@@
 \end{align}
+Observing five initial packets, $\text{Size}_\text{init}$ is the packet size feature-frequency vector for only those five packets.
+Similarly, $\text{Ival}_\text{init}$ is the feature-frequency vector for the first four inter-packet intervals.
 The matrix of ``predictables'' would be:
@@
   X_\text{p} = \bracket{ ~
-    \text{Size}_{\text{rest}} ~
-    \text{Ival}_{\text{rest}} ~
+    \text{Size}_\text{rest} ~
+    \text{Ival}_\text{rest} ~
     \text{Pkts} ~
@@
 \end{align}
-We use the packet sizes and inter-packet intervals of the initial packets, together with their \caps{IP} type and port numbers to predict the distribution of packets sizes and inter-packet intervals for the rest of the flow, as well as predicting a distribution of possible of packet counts.
-In order to do this prediction, subspace factorization must be done with the observable and predictable matrices separated.
-That is, subspace factorization must be done on the matrix $[X_\text{o}~X_\text{p}]$.
+The matrices $\text{Size}_\text{rest}$ and $\text{Ival}_\text{rest}$ are the analogous  matrices for the rest of the packets in the flows.
+
+From the observed data, $X_\text{o}$, we find the best fitting weight matrix, and use the weights together with our recovered model parameters to estimate probability distributions of features for each flow, using the hypothesized model.
+This technique is fundamentally similar to fitting a regression line to noisy two-dimensional training data and then using the $x$-coordinates of data points to linearly predict the $y$-coordinates.
+The main differences are the vastly higher number of dimensions and more complex structure we assume for our data.
+
+% We use the packet sizes and inter-packet intervals of the initial packets, together with their \caps{IP} type and port numbers to predict the distribution of packets sizes and inter-packet intervals for the rest of the flow, as well as predicting a distribution of possible of packet counts.
+% In order to do this prediction, subspace factorization must be done with the observable and predictable matrices separated.
+% That is, subspace factorization must be done on the matrix $[X_\text{o}~X_\text{p}]$.
 % However, we perform our subspace segmentation step on the combined data as in \Equation{X}.
@@
 
-Our technique is fundamentally similar to fitting a regression line to noisy two-dimensional training data and then using the $x$-coordinates of test data points to linearly predict the $y$-coordinates.
-The main differences are the vastly higher number of dimensions and more complex structure we assume for our data.
-From the training data, we recover our estimated version of the $P$ matrix.
-We split this into its observable and predictable parts: $P_\text{o}$ and $P_\text{p}$.
-Given an observation matrix, $X^*_\text{o}$, to perform prediction, we find the nonnegative matrix of weights, $W_*$, minimizing the squared Frobenius error:
+\newfootnote{\separationnote}{%
+We use a combined feature matrix for subspace segmentation, but separated observable and predictable features for subspace factorization.
+This is partly because segmentation appears to work better using combined features, but also so that we can reuse the same expensive segmentation computation when performing prediction with different numbers of initial packets.
+}
+
+Let $P^* = \bracketx{P^*_\text{o}~P^*_\text{p}}$ be the recovered model parameter matrix from ~\Section{model-recovery}, with separated observable and predictable features.\separationnote
+From an observation matrix, $X_\text{o}$, we estimate the matrix of weights by minimizing the squared Frobenius error:
 \begin{align}
-  \norm{X^*_\text{o} - W^* P_\text{o}}^2_\text{frob}.
+  W^* = \text{argmin}_W {\norm{X_\text{o} - W P^*_\text{o}}^2_\text{frob}}.
 \end{align}
-This is equivalent to minimizing the square error for each observation individually, but optimizing \emph{en masse} allows fast non-negative least squares algorithms to be used~\cite{Benthem04,Kim08:block-pivot}.
-The predicted behavior for these flows is simply the weight matrix times the predictable feature matrix:
+Note that since the weights must be nonnegative, a nonnegative least squares algorithm must be used~\cite{Benthem04,Kim08:block-pivot}.
+From here, we can estimate the underlying feature distributions for the flows:
 \begin{align}
-  X^*_\text{p} = W^* P_\text{p}.
+  Q^* = W^*P.
 \end{align}
-The rows of $X^*_\text{p}$ are our predictions about the behavior of flows observed in $X^*_\text{o}$.
-Each row contains a packet size distribution, an inter-packet interval distribution and a distribution of packet counts.
-
-% With our hypothetical model parameters, $p_j$, extracted from training data, for each test observation, $q^*_\text{o}$, we find nonnegative weights, $w^*_j$, that minimize the squared error
-% \begin{align}\textstyle
-%   \norm{q^*_\text{o}-\sum {w^*_j p_j}}^2.
-% \end{align}
-% The predicted behavior of the flow is simply
+In particular, the ``predictable'' portion, $Q^*_\text{p}$, contains predictions of packet size distribution, inter-packet interval distribution and distribution of packet counts for each flow.
+To evaluate the quality of these predictions, we compare the distributions in $Q^*_\text{p}$ to the matrix, $X_\text{p}$, of actual behaviors of flows.
+The next section explores how to preform this evaluation and how our technique performs compared with other possible predictors.
 

### 2009-07-31 paper.tex
...

@@
 When compared with a specific outcome, such a prediction cannot be said to be unequivocally right or wrong.
-How then can we measure the quality of a predicted distribution of outcomes as compared to a single outcome?
-This subtle and difficult issue is central to our evaluation methodology.
+% How then can we measure the quality of a predicted distribution of outcomes as compared to a single outcome?
+% This subtle and difficult issue is central to our evaluation methodology.
 % In what follows, we have performed prediction from observing only five initial packets from each flow.
@@
 
-To evaluate the quality of predicted distributions as compared with a sample that may or may not come from that distribution, we use the single-sample Kolmogorov-Smirnov (K-S) ``goodness-of-fit'' test~\cite{Feller68}.
-The K-S test compares the cumulative distribution function (\caps{CDF}) of a theoretical distribution with the empirical cumulative distribution function of a sample of values.
-The test uses the largest absolute difference between the two \caps{CDF}s as a statistic to measure how plausible it is that the sample was drawn from the theoretical distribution.
+To test the quality of a predicted distribution as compared with a sample of values which may or may not come from that distribution, we use the single-sample Kolmogorov-Smirnov (K-S) ``goodness-of-fit'' test~\cite{Feller68}.
+The K-S test compares the cumulative distribution function (\caps{CDF}) of a theoretical distribution with the empirical cumulative distribution function of the sample.
+The test uses the largest absolute difference between the \caps{CDF}s as a statistic to measure how plausible it is that the sample was drawn from the hypothesized distribution.
 The K-S statistic can be transformed into a p-value:

### 2009-07-31 paper.tex
...

@@
 \end{center}
-\vspace{-1.5em}
+\vspace{-1.6em}
 \end{figure*}
@@
 The test uses the largest absolute difference between the \caps{CDF}s as a statistic to measure how plausible it is that the sample was drawn from the hypothesized distribution.
-The K-S statistic can be transformed into a p-value:
-the probability of obtaining a result at least as extreme as the observed result, assuming that the sample really did come from the theoretical distribution.
+The K-S statistic is transformed into a p-value:
+the probability of obtaining a result as extreme as observed, assuming the sample really came from the hypothesized distribution.
 
-\newfootnote{\ksidealnote}{If the K-S test were ideal, this pragmatic baseline would in fact be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
+\newfootnote{\ksidealnote}{If the K-S test were perfect, the pragmatic baseline would be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
 
-We evaluate the quality of our packet size and inter-packet interval distribution predictions by exampining the distribution of K-S p-values for the predicted distribution versus the observed sample.
-In theory, if the predictions are all correct, the p-values should be uniformly distributed on the interval [0,1].
+We evaluate the quality of our packet size and inter-packet interval distribution predictions by examining the distribution of K-S p-values for the predicted distribution versus the observed sample.
+In theory, if the predictions are correct, the p-values should be uniformly distributed on the interval [0,1].
 In practice, however, the K-S test is not perfect, and its p-values are not uniform.
@@
 For comparison, we also show the results for two simplistic prediction techniques.
-The first alternative---``mean''---always predicts the average behavior of the entire training trace.
+The first alternative, ``mean,'' always predicts the average behavior of the entire training trace.
 This does not perform as badly as one might think it would:
-always guessing the mean is not an abysmal failure.
-The second alternative---``self''---predicts the behavior that has already been seen.
+always guessing the average is not an abysmal failure.
+The second alternative, ``self,'' predicts the behavior that has already been seen.
 % This is the best guess one can make about a flow, without having been trained on any other flows:
@@
 
-A na\"ive point-measure of the error of a prediction technique is the rate at which it passes the K-S test at a predetermined significance level, such as 1\% or 5\%, as compred to the ideal.
+A na\"ive point-measure of the error of these predictions is the rate at which they fail the K-S test at a fixed significance level, such as 1\% or 5\%, as compared to the ideal.
 This is equivalent to measuring the distance above the ideal \caps{CDF} at the 0.01 or 0.05 p-value mark.
@@
 they are chosen so as to incorrectly reject the null only 1\% or 5\% of the time.
-Thus they only reject very bad predictions.
+Thus, they only reject the worst possible predictions.
 The mean predictor makes a perfect example:
-by guessing the average behavior it actually makes relatively few really bad predictions but makes far more poor to mediocre predictions than it should, and almost no good predictions.
+by guessing the average behavior it actually makes relatively few really bad predictions but makes far more poor and mediocre predictions than it should, and very few good predictions.
 Our predictor, on the other hand, maintains a fairly steady error rate somewhat above the ideal.

### 2009-07-31 paper.tex
...

@@
 \end{center}
-\vspace{-1.6em}
+\vspace{-1.5em}
 \end{figure*}
@@
 by guessing the average behavior it actually makes relatively few really bad predictions but makes far more poor and mediocre predictions than it should, and very few good predictions.
-Our predictor, on the other hand, maintains a fairly steady error rate somewhat above the ideal.
+Our predictor, on the other hand, maintains a fairly steady error rate somewhat above the ideal, and in particular makes as many near-perfect predictions as the ideal.
 
-As a better point-measure of the quality of predictors, we propose the maximum vertical distance between its p-value \caps{CDF} and the ideal.
+\begin{figure*}[!t]
+\vspace{-1em}
+\begin{center}
+\subfloat[CDF of accuracy p-values]{\fig{packets-accuracy}%
+\includegraphics[width=2.351in]{pkts_accuracy}}
+\subfloat[CDF of prediction entropy values]{\fig{packets-entropy}%
+\includegraphics[width=2.351in]{pkts_entropy}}
+\subfloat[CDF of reduction of uncertainty]{\fig{packets-uncertainty}%
+\includegraphics[width=2.351in]{pkts_certainty}}
+\caption{%
+CDFs displaying the accuracy, uncertainty reduction, and entropy of our packet count predictions.
+Ideal behavior in \Figure{packets-accuracy} is as close to the diagonal as possible.
+For \Figure{packets-entropy}, lower entropy is better.
+the vertical lines indicate the entropy of the mean behavior for each trace;
+prediction entropy must be compared to this.
+In \Figure{packets-uncertainty}, the greater the reduction in uncertainty, the better.
+}
+\end{center}
+\vspace{-2em}
+\end{figure*}
+
+As a better point-measure of the quality of a predictor, we propose the maximum vertical distance between its p-value \caps{CDF} and the ideal.
 This is equivalent to choosing the worst possible predetermined significance level for each predictor and evaluating it at that level.
@@
 \Figure{success-rates} shows the success rates of various predictions across the six data sets.
+Our method predicts size and interval behaviors with accuracy between 70\% and 90\% across all data sets, while the other methods remain well below 50\% few exceptions.
 
-\begin{figure}[!b]
+\begin{figure}[b]
 \vspace{-1em}
@@
 
-% \begin{table}[b]
-% \begin{center}
-% \small
-% \begin{tabular}{|c|c|c|c|c|c|c|}
-% \hline
-% &
-% \multicolumn{2}{c|}{\textbf{Predicted}} &
-% \multicolumn{2}{c|}{\textbf{Self}} &
-% \multicolumn{2}{c|}{\textbf{Mean}} \\
-% \hline
-% \textbf{Trace} &
-% \textbf{size} & \textbf{ival} &
-% \textbf{size} & \textbf{ival} &
-% \textbf{size} & \textbf{ival} \\
-% \hline
-% \input{pred_stats}
-% \hline
-% \end{tabular}
-% \caption{Error rates for size and interval predictions.}
-% \tab{error-rates}
-% \end{center}
-% \vspace{-1em}
-% \end{table}
-
-\begin{figure*}[t]
-\vspace{-1em}
-\begin{center}
-\subfloat[Packet count: accuracy]{\fig{packets-accuracy}%
-\includegraphics[width=2.351in]{pkts_accuracy}}
-\subfloat[Packet count: uncertainty]{\fig{packets-uncertainty}%
-\includegraphics[width=2.351in]{pkts_certainty}}
-\subfloat[Packet count: entropy]{\fig{packets-entropy}%
-\includegraphics[width=2.351in]{pkts_entropy}}
-\caption{Packet count predictions: accuracy, uncertainty reduction, and distribution entropy.}
-\end{center}
-\vspace{-2em}
-\end{figure*}
-
 Next we turn to the evaluation of packet counts predictions.
 Here we cannot use the K-S test:
-although we have a predicted distribution of possible packet counts, we have only one value to compare it to---the actual number of packets.
-The K-S test is useless for samples of fewer than three values---it cannot invalidate any outcome.
-Instead we use the predicted distributions themselves to yield p-values for each packet count.
-We compute how ``extreme'' each observed packet count value is according to the predicted distribution of packet count values:
-that is, what fraction of the distribution is as large or larger than the observed value.
+although we have a predicted distribution of possible packet counts, we have only one value to compare it to---the observed number of packets.
+Instead we use the predicted distributions themselves to produce p-values for observed packet counts.
+We compute how ``extreme'' each observed packet count value is in the predicted distribution of values:
+what fraction of the distribution is as large or larger than the observed value.
 If the predictions are accurate, then these ``p-values'' should be uniformly distributed.
 
+\Figure{packets-accuracy} shows the \caps{CDF} of p-values of observed packet counts versus our predicted distributions.
+While the \caps{CDF}s do not remain perfectly straight along the diagonal, the general tendency is for our predictions to remain fairly accurate.
+There is some tendency for the observations to fall at the high and low ends of the predicted distributions, indicating that our predictions are somewhat too ``narrow.''
+Nevertheless, the general nearness to the diagonal shows that, on the whole, our packet predictions remain fairly accurate.
+
 Accuracy alone is insufficient, however:
-guessing the mean distribution of packet counts yields a perfectly accurate predictor according to this criterion.
-What we want, however, is to remain accurate but become more certain about the packet count.
-We can measure this certainty by computing the information-theoretic entropy of each prediction.
+guessing the mean distribution of packet counts yields a perfectly accurate predictor by this criterion.
+What we want, however, is to remain accurate but become more precise in our predictions of packet count distributions.
+We can measure this precision by computing the information-theoretic entropy of each distribution.
 We want our accuracy to remain intact, but the entropy of our predictions to go down as much as possible.
-A related measure is the reduction in uncertainty:
-the exponential of reduction in entropy from the mean predictor.
+
+\Figure{packets-entropy} shows the \caps{CDF}s of entropy values for our predictions across all six data sets.
+The entropy of the average behavior of each set is indicated with a vertical line.
+Entropy of predictions must be evaluated as compared with this.
+In some cases, the entropy of our prediction is greater than the entropy of the average behavior.
+In such cases, we can simply use the average behavior as a prediction instead.
+Most of the time, however, we are gaining a great deal of information about what the actual packet count will be:
+entropy is reduced to less than two bits or less about half of the time.
+
+A closely related measure to entropy is the reduction in uncertainty.
+The uncertainty of a prediction is two raised to the power of its entropy:
+the number of possible choices the bits of entropy can encode.
+The reduction in uncertainty is the fraction of the mean behavior's uncertainty that remains when we use our predictor.
 This is the fraction of the original ``number'' of possible values that remain after our prediction as compared to before.
-We put ``number'' in quotes because even though entropy is measured in bits, it doesn't count enumerable discrete values like a byte value would.
-
-\Figure{packets-accuracy} shows the \caps{CDF} of p-values of observed packet counts versus our predicted distributions.
-While some of the \caps{CDF}s do not remain perfectly straight along the diagonal, overall the tendency is for our predictions to remain accurate.
-There is some tendency for the observations to fall at the high and low ends of our predictions, indicating that we are in some cases making predictions that are too narrow.
-Still the general adherence to the diagonal shows that oh the whole, the packet predictions remain fairly accurate.
-\Figure{packets-uncertainty} shows that while we are not losing much accuracy, we are gaining a great deal of certainty about what the actual packet count will be.
-Most of the time, 50\% of the time, we discard around 80\% of the original uncertainty.
-\Figure{packets-entropy} complements this figure, showing what the actual distribution of entropy values for the six data sets.
+\Figure{packets-uncertainty} shows that across the six data sets, half of the time we discard between 75\% and 90\% of the original uncertainty.
 

### 2009-07-31 paper.tex
...

@@
 \newfootnote{\separationnote}{%
-We use a combined feature matrix for subspace segmentation, but separated observable and predictable features for subspace factorization.
+We use an unseparated feature matrix for subspace segmentation, but separated observable and predictable features for subspace factorization.
 This is partly because segmentation appears to work better using combined features, but also so that we can reuse the same expensive segmentation computation when performing prediction with different numbers of initial packets.
@@
 \Figure{success-rates} shows the success rates of various predictions across the six data sets.
-Our method predicts size and interval behaviors with accuracy between 70\% and 90\% across all data sets, while the other methods remain well below 50\% few exceptions.
+Our method predicts size and interval behaviors with accuracy between 70\% and 90\% across all data sets, while the other methods remain well below 50\% with few exceptions.
 
@@
 
-Our results show that we can predict the packet size distribution and inter-packet interval distribution of flows from only observing a handful of initial packets of each flow.
+Our results show that we can predict the packet size distribution and inter-packet interval distribution of flows from only a handful of initial packets of each flow.
 Specifically, we have used five initial packets to predict these distributions with between 70\% and 90\% accuracy across a wide variety of network traces.
@@
 Immediate applications of this ability to accurately predict individual flow behaviors abound.
-Quality of service (\caps{QoS}) schemes attempting to provide high-quality service guarantees for streaming voice and video wireless through wireless networks could use such predictions to know---almost presciently---how much data flows will transmit and how often.
+Quality of service (\caps{QoS}) schemes attempting to provide high-quality service guarantees for streaming voice and video wireless through wireless networks could use such predictions to know, almost presciently, how much data flows will transmit and how often.
+Routers could use information about data rates from these predictions to load-balance across multiple possible routes towards their ultimate destinations.
 

### 2009-07-31 paper.tex
...

@@
 
-Our results show that we can predict the packet size distribution and inter-packet interval distribution of flows from only a handful of initial packets of each flow.
+Our results show that we can predict the packet size distribution, inter-packet interval distribution, and packet count distribution of flows from only a handful of initial packets of each flow.
 Specifically, we have used five initial packets to predict these distributions with between 70\% and 90\% accuracy across a wide variety of network traces.
-Moreover, we can reduce our uncertainty about the number of packets that flows will have by 80\% or more, about half of the time.
+Moreover, about half of the time, we can reduce our uncertainty about the number of packets that flows will have by 80\% or more.
 Immediate applications of this ability to accurately predict individual flow behaviors abound.
-Quality of service (\caps{QoS}) schemes attempting to provide high-quality service guarantees for streaming voice and video wireless through wireless networks could use such predictions to know, almost presciently, how much data flows will transmit and how often.
+Quality of service (\caps{QoS}) schemes attempting to provide high-quality service guarantees for streaming voice and video wireless through wireless networks could use such predictions to know, almost presciently, how much data flows are likely to transmit and how often.
 Routers could use information about data rates from these predictions to load-balance across multiple possible routes towards their ultimate destinations.
 
-The real importance of this work has little to do with these potential practical applications.
-The ability to predict flow behavior is merely a pleasant side effect of the real breakthrough:
+While these potential applications are useful, the real significance of this work lies elsewhere.
+Predicting flow behavior is merely a pleasant side effect of the real breakthrough:
 a detailed statistical model for network-wide flow behaviors.
-The fundamental concept underlying this work is that there are a relatively small collection of ``basic behaviors'' that generate the vast majority of the flow behaviors that we actually see in any particular trace.
+The fundamental concept underlying this work is that there are a relatively small collection of ``basic behaviors'' that generate the vast majority of the flow behaviors found in any particular trace.
 These behaviors are associated with distributions of various flow features:
 packet sizes, inter-packet intervals, \caps{IP} protocol, port numbers and packet counts.
-these features are drawn from these distributions to ``construct'' actual flows.
-We have treated the features as though they are sampled independently, but no doubt they are not really independent at all.
+These features are drawn from these distributions to create actual flows.
+% We have treated the features as though they are sampled independently, but no doubt they are not really independent at all.
 

### 2009-07-31 paper.tex
...

@@
 
-\newfootnote{\ksidealnote}{If the K-S test were perfect, the pragmatic baseline would be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
+\newfootnote{\ksidealnote}{If the K-S test were perfect, the pragmatic ideal baseline would be uniform. The fact that it is not demonstrates the imperfection of the K-S test.}
 
@@
 
-Part of the reason the model of flow generation as mixtures of a basic set of behaviors is not obvious and has not previously been noted is likely that with convex linear combinations, so many different combinations can be created.
-Moreover, the analytical tools to extract this structure have only existed for a few years.
+Given that the statistical model we have proposed for flow behavior is fairly simple---and finite mixture modeling is a well established field that has been studied since 1894~\cite{McLachlan00}.
+Traditional approaches, however, have only applied to parametric probability distributions.
+The distributions found in network traffic, however, are very distinctly non-parametric.
+They are lumpy, unruly and generally quite ugly.
+This is due largely to the fact that behaviors found in computer networks are not the by-products of smooth, natural phenomena, but rather of man-made artificial systems, following arbitrary, complex and capricious rules.
+
+The numerical techniques needed to tackle non-parametric mixture modeling have only very recently become available.
 In our model recovery procedure, we use a subspace segmentation algorithm published in 2007~\cite{Ma07}, two \caps{NMF} algorithms published in 2001 and 2008, respectively~\cite{Lee01,Kim08:anls}, and a low-rank \caps{NMF} initialization technique published for the first time here.
-And these techniques are still quite immature:
+All of these techniques are still quite immature:
+New \caps{NMF} algorithms have been regularly proposed since Lee and Seung's seminal paper in 2001.
+Their algorithms are still the most well-known, and often the most reliable despite their various shortcomings~\cite{Langville06}.
 they will only improve over time, yielding even better recovery of the model parameters.
 Such improvements will immediately improve the performance of the prediction methods employed in this paper.
+
+% Part of the reason the model of flow generation as mixtures of a basic set of behaviors is not obvious and has not previously been noted is likely that with convex linear combinations, so many different combinations can be created.
+% Moreover, the analytical tools to extract this structure have only existed for a few years.
+% In our model recovery procedure, we use a subspace segmentation algorithm published in 2007~\cite{Ma07}, two \caps{NMF} algorithms published in 2001 and 2008, respectively~\cite{Lee01,Kim08:anls}, and a low-rank \caps{NMF} initialization technique published for the first time here.
+% And these techniques are still quite immature:
+% they will only improve over time, yielding even better recovery of the model parameters.
+% Such improvements will immediately improve the performance of the prediction methods employed in this paper.
 
@@
 % TODO: mention the two dimension structure assumption.
-% TODO: mention the five packets in the results somewhere.
+
+\vfill
+\pagebreak
 

### 2009-07-31 paper.tex
...

@@
 
-Our results show that we can predict the packet size distribution, inter-packet interval distribution, and packet count distribution of flows from only a handful of initial packets of each flow.
+The results demonstrate that we can predict the packet size distribution, inter-packet interval distribution, and packet count distribution of flows from only a handful of initial packets of each flow.
 Specifically, we have used five initial packets to predict these distributions with between 70\% and 90\% accuracy across a wide variety of network traces.
@@
 Their algorithms are still the most well-known, and often the most reliable despite their various shortcomings~\cite{Langville06}.
-they will only improve over time, yielding even better recovery of the model parameters.
+These algorithms will only improve over time, yielding better recovery of model parameters.
 Such improvements will immediately improve the performance of the prediction methods employed in this paper.

### 2009-07-31 paper.tex
...

@@
 The experimental methodology for testing this hypothesis is presented in \Section{methodology} while the results follow in \Section{results}.
-\Section{discussion} discusses the meaning and impact of these results and
-\Section{related-work} gives context for this research in relation prior work.
+\Section{discussion} discusses the meaning and impact of these results and gives context for this research in relation prior work.
 We conclude with final remarks in \Section{conclusions}.
@@
 
-\section{Discussion}\sec{discussion}
+\section{Discussion \& Related Work}\sec{discussion}\sec{related-work}
 
@@
 These features are drawn from these distributions to create actual flows.
-% We have treated the features as though they are sampled independently, but no doubt they are not really independent at all.
+We have treated the features as though they are sampled independently, but this is merely a simplification that allows us to estimate model parameters with computational techniques available today.
+The fact that this estimation provides a model which can reasonably well predict the behaviors of flows demonstrates that the concept of modeling flows as a mixture of basic behaviors is viable, even if the details of how we represent flow features remains open to improvement.
 
-Given that the statistical model we have proposed for flow behavior is fairly simple---and finite mixture modeling is a well established field that has been studied since 1894~\cite{McLachlan00}.
-Traditional approaches, however, have only applied to parametric probability distributions.
+Given that the statistical model we have proposed for flow behavior is fairly simple---and finite mixture modeling is a well established field that has been studied since 1894~\cite{McLachlan00}---one may very well ask why this basic linear structure of network traffic has not been discovered before.
+Part of the reason is that traditional approaches to mixture modeling have been applicable to parametric probability distributions.
 The distributions found in network traffic, however, are very distinctly non-parametric.
@@
 The recovered model used in this paper is very close to providing a usable model for traffic generation already.
-The largest missing piece is a model for realistic weight matrices.
+The largest missing piece is a model for generating realistic weight matrices.
+With realistic basic behavior matrices extracted from traces and a means of generating realistic weight matrices, we have immediately a model for generating realistic traffic traces.
+Earlier research has show that it is sufficient for simulation purposes to generate flows with realistic packet size distributions, inter-packet interval distributions, and packet counts by repeatedly sampling packets from these distributions~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
 If such applications based on this flow behavior model or other similar ones prove as fruitful as the application of flow behavior prediction, that will provide further evidence of the effectiveness and validity of our proposed model.
@@
 % TODO: mention the two dimension structure assumption.
-
-\vfill
-\pagebreak
-
-\section{Related Work}\sec{related-work}
 

### 2009-07-31 paper.tex
...

@@
 
+We are not the first researchers to apply linear techniques to the behavior of network traffic.
+Lakhina~\emph{et~al.}~\cite{Lakhina04} apply principal component analysis (\caps{PCA})---which is computationally equivalent to \caps{SVD}---to the aggregate packet count traversing between ingress and egress points in Internet transit networks over time.
+Their purpose for this application is to detect network traffic anomalies, but in the process, they introduce the concept of ``eigenflows'':
+the principal components (\emph{c.f.} singular vectors) of aggregate traffic time series.
+De Oliveira~\emph{et~al.}~\cite{DeOliveira06} further the application to \caps{PCA} to ingress-egress aggregate traffic volume, which the literature terms ``origin-destination flows,'' by classifying the principal components into three classes:
+periodic components, peaked components, and noise components.
+The same sorts of analysis have be extended by several other researchers~\cite{Babiarz06,Juva08}.
+While some of the techniques are similar, such as \caps{SVD} and \caps{PCA}, and the terminology seems suggests similarity---all talk about ``flows''---the reality is that this research is a significant departure.
+These earlier approaches analyze and forecast aggregate traffic volume in transit networks, while our research predicts the behaviors of individual flows, in the sense of a series of packets sharing a ``5-tuple,'' rather than the transit network ``origin-destination'' pair sense.
+While understanding aggregate traffic behavior in core networks is certainly useful, we believe that an even broader application exists for the detailed individual flow behavior model we propose and provide evidence for here.
+
 % Part of the reason the model of flow generation as mixtures of a basic set of behaviors is not obvious and has not previously been noted is likely that with convex linear combinations, so many different combinations can be created.
@@
 
-It is our hope and belief that this model of flow behavior will provide future insights into other aspects of network traffic modeling:
+It is our hope and belief that our model of flow behavior will provide future insights into other aspects of network traffic modeling:
 in particular traffic classification and realistic workload generation.
@@
 With realistic basic behavior matrices extracted from traces and a means of generating realistic weight matrices, we have immediately a model for generating realistic traffic traces.
-Earlier research has show that it is sufficient for simulation purposes to generate flows with realistic packet size distributions, inter-packet interval distributions, and packet counts by repeatedly sampling packets from these distributions~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
-If such applications based on this flow behavior model or other similar ones prove as fruitful as the application of flow behavior prediction, that will provide further evidence of the effectiveness and validity of our proposed model.
+Earlier research has show that it is sufficient for simulation purposes to generate flows with realistic packet size distributions, inter-packet interval distributions, and packet counts, repeatedly sampling packets from these distributions~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
+If such applications based on this flow behavior model or other similar ones prove as fruitful as the present application to flow behavior prediction, those results will provide further evidence of the effectiveness and validity of our proposed model.
 

### 2009-07-31 paper.tex
...

@@
 These earlier approaches analyze and forecast aggregate traffic volume in transit networks, while our research predicts the behaviors of individual flows, in the sense of a series of packets sharing a ``5-tuple,'' rather than the transit network ``origin-destination'' pair sense.
-While understanding aggregate traffic behavior in core networks is certainly useful, we believe that an even broader application exists for the detailed individual flow behavior model we propose and provide evidence for here.
+While understanding aggregate traffic behavior in core networks is certainly useful, we believe that even broader applications exist for the detailed individual flow behavior model we have proposed.
 

### 2009-07-31 paper.tex
...

@@
 \end{center}
-\vspace{-2em}
+\vspace{-2.5em}
 \end{table*}
@@
 Now we must determine the vertices, $p_j$, which generate these structures.
+The first question is how many ``real'' dimensions each structure has.
+Various methods exist for trying to determine this.
+In \Figure{svd} it appears that most of our structures are nearly two-dimensional.
+Despite attempting many more sophisticated techniques for automatically determining structure dimensionality, we have found that best prediction quality comes from simply assuming that each structure is two-dimensional.
 We can use \caps{SVD} to determine the most prominent linear components of each structure, but \caps{SVD} will give us vectors of mixed sign.\svdnnnote

### 2009-07-31 paper.tex
...

@@
 
+\section{Conclusions}\sec{conclusions}
+
+From a relatively simple observation about apparent linear structure in the feature-frequency of network traffic traces, we have proposed that flow behaviors traces follow a specific type of statistical model:
+a non-parametric finite mixture model.
+In this model, each flow's behavior distribution is a convex linear combination of a few ``basic behaviors,'' and the features of the flow are drawn from the resulting hybrid distribution of behaviors.
+We are able to use recently developed data mining tools to recover estimates of the model's parameter matrix from training trace data.
+Using the recovered trace parameters, we can perform prediction on flow behaviors to predict their behavior from observing only a few initial packets of each flow.
+In our experiments, we observe only five initial packets of each flow and succeed in predicting these distribution of packet sizes and inter-packet intervals with between 70\% and 90\% accuracy across a wide variety of network traces.
+Also, half of the time, we can reduce our uncertainty about the number of packets that flows will have by 80\% or more, giving very important early information about the likely duration of each flow.
+Combining these distribution predictions, one can readily predict distributions for the duration and total data volume of each flow.
+
+Despite the obvious immediate practical applications of the prediction techniques demonstrated here, the real importance of this work lies in the theoretical underpinnings.
+The proposed model of flow behavior as a finite mixture model is the first detailed statistical model of universal flow behavior.
+Such a model stands to have great impact on the networking community.
 It is our hope and belief that our model of flow behavior will provide future insights into other aspects of network traffic modeling:
@@
 
-% TODO: mention the two dimension structure assumption.
+% The next obvious application is to use this structure to perform better traffic analysis and classification, but ultimately the most powerful use of such modeling is to generate realistic workload for network simulations.
+% As noted by other research~\cite{Kaprinski07:realism,Karpinski07:cbr-failure}, the lack of realistic models for wireless workload
 
-\section{Conclusions}\sec{conclusions}
+\vfill
+\pagebreak
 
